---
title: "SOC6707: Intermediate Data Analysis"
author: "Monica Alexander"
date: "Week 2: Exploratory Data Analysis and Data Visualization"
output: 
  beamer_presentation:
    slide_level: 2
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, size = '\tiny')
```

## Overview

What we will cover today:

- What is EDA and why do we do it?
- Steps of EDA
- Data visualization
- Key types of graphs
    + another summary measure: correlation coefficient
- Doing EDA and data viz in R (intro to `ggplot`)

(Exploratory Data Analysis = EDA)

# Exploratory Data Analysis (EDA)

## What is EDA and why do we do it?

Before we even do any sort of statistical inference, we need to understand the main characteristics of our dataset. 

- Helps to identify any potential issues or surprising things about our data
- Helps to check / explore / refine research questions

## What is EDA and why do we do it?

EDA is all about asking: 

- What types of variables do we have?
- Do we have a complete dataset, or do we have missing data or observations?
- If we have missing data, is it missing equally across observations of different types or concentrated in particular groups?
- Are there any obvious outliers or strange data points?
- What do the data 'look' like? 
    + summary measures, measures of centrality, spread
    + Visualizing the data through plots and tables
    
## Steps of EDA

1. Become familiar with size of data set (number of observations and variables available)
2. What kinds of variables are available
3. For the variables that I'm interested in, are there any missing values or other issues?
4. What does the distribution/frequency of observations look like for the variables I'm interested in? (summary measures, tables and graphs)

## Example: TTC subway delays in 2019

- Data on TTC subway delay times by station and day available from the Open Data Toronto website: https://open.toronto.ca/
- Accessed through using the `opendatatoronto` R package: https://sharlagelfand.github.io/opendatatoronto/

```{r, echo = FALSE}
library(tidyverse)
library(here)
delay_2019 <- read_csv(here("data/ttc_delays_2019.csv"))
```

\includegraphics{../fig/ttc.jpg}


## Get familiar with dataset

\tiny
```{r}
delay_2019
```

## Get familiar with dataset

Dimensions (number of rows x number of columns)

\tiny

```{r}
dim(delay_2019)
```

Variable names

\tiny

```{r}
colnames(delay_2019)
```

## The `summary` function is useful for a quick overview

\tiny
```{r}
summary(delay_2019)
```

## Research question?

- What are some good potential research questions with this dataset?

## Sanity checks

We need to check variables should be what they say they are. If they aren't, the natural next question is to what to do with issues (recode? remove?)

E.g. check days of week make sense with the `unique` function

\tiny

```{r}
delay_2019 %>% 
  select(day) %>% 
  unique()
```

## Sanity checks

Check lines: oh no. some issues here. Some have obvious recodes, others, not so much. 
\tiny
```{r}
delay_2019 %>% 
  select(line) %>% 
  unique() %>% 
  pull() # turn into a vector for better display
```
## Data issues

How bad is the mislabeling of lines? look at frequency of cases

NOTE! New very important function: `group_by` 

\tiny

```{r}
delay_2019 %>% 
  group_by(line) %>% # group by line label
  tally() %>% # count the number of occurrences
  arrange(-n) # arrange in descending order
```
## Missing values

\tiny

```{r}
delay_2019 %>% 
  summarise_all(.funs = funs(sum(is.na(.))))
```


<!-- ## Or use the `skimr` package -->

<!-- \tiny  -->

<!-- ```{r, eval = FALSE} -->
<!-- library(skimr) # install using install.packages("skimr") -->
<!-- skim(delay_2019) -->
<!-- ``` -->

<!-- \normalsize -->
<!-- (Show this in lecture) -->

```{r, echo = FALSE}
library(janitor)
delay_2019 <- delay_2019 %>% distinct()
delay_2019 <- delay_2019 %>% filter(line %in% c("BD", "YU", "SHP", "SRT")) 
```


## Summary statistics

Most interested in delay minutes, which is the `min_delay` variable

\tiny

```{r}
delay_2019 %>% 
  summarize(n_obs = n(), 
            mean_delay = mean(min_delay),
            median_delay = median(min_delay),
            range_delay = max(min_delay) - min(min_delay),
            iqr_delay = IQR(min_delay))
```

## Summary statistics

Probably more interesting to do these summaries by line (**stratify** by line); easy extension with the `group_by` function

\tiny

```{r}
delay_2019 %>% 
  group_by(line) %>% 
  summarize(n_obs = n(), 
            mean_delay = mean(min_delay),
            median_delay = median(min_delay),
            range_delay = max(min_delay) - min(min_delay),
            iqr_delay = IQR(min_delay))
```


## Summaries

Could also stratify by reason for delay

\tiny

```{r}
delay_2019 %>% 
  group_by(code_desc) %>% 
  summarize(n_obs = n(), 
            mean_delay = mean(min_delay),
            median_delay = median(min_delay),
            range_delay = max(min_delay) - min(min_delay),
            iqr_delay = IQR(min_delay)) %>% 
  arrange(-n_obs)
```

## Summaries

Arrange by mean delay time

\tiny

```{r}
delay_2019 %>% 
  group_by(code_desc) %>% 
  summarize(n_obs = n(), 
            mean_delay = mean(min_delay),
            median_delay = median(min_delay),
            range_delay = max(min_delay) - min(min_delay),
            iqr_delay = IQR(min_delay)) %>% 
  arrange(-mean_delay) 
```
## EDA: summary so far

- There's no one checklist of things to looks at, depends on your data and research question
- Get familiar with your dataset
- Check for missing values, and that existing values make sense
- Summary statistics depend on your research question of interest
    + stratifying (`group_by`) by important characteristics often useful 


# Data visualization

## Plot your data!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!

- We started to compute some summary statistics above, and showed how summaries can be calculated by group and arranged in different ways to get a sense of differences across groups
- However, graphing/plotting your data is usually the best way to visualize patterns, trends, outliers, issues and other surprising points
- The most appropriate types of graph for your data depends on:
    + the type of variable you are interested in (quantitative or qualitative/categorical)
    + your research questions


## Plot your data!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!

- Before you start to do any statistical analysis, you should always plot your data
- Data visualization is a key part of EDA and essential in understanding the assumptions and outcomes of your eventual statistical analysis

# Important types of graphs

## Important types of graphs

- Histograms
- Bar charts
- Boxplots
- Line plots
- Scatter plots 


## Example datasets used here

1.  TTC subway delays (from above)
2. Country-level indicators, 2009-2017
    + Uploaded onto Quercus
    + TFR = total fertility rate
    + GDP = gross domestics product
    + dataset also has life expectancy (females), child mortality, maternal mortality
    
    
## Histograms

Shows the distribution of a **quantitative** variable

- Histograms show the frequency (count) of observations by value 
- The range of values of a variables is divided into intervals ('bins') and then the number of observations in each bin is tabulated
- A histogram shows the count of observations in each bin with a rectangle of height equal to the count
- The x axis is the value bins, the y axis is the count/frequency (or proportion)

## 

```{r, echo = FALSE}
country_ind <- read_csv("../../data/country_indicators.csv")
country_ind %>% 
  filter(year==2017) %>% 
  ggplot(aes(life_expectancy)) + 
  geom_histogram(binwidth = 2, fill = "firebrick", color = "white") + 
  theme_bw(base_size = 14) + 
  xlab("life expectancy") + 
  labs(title = "Female life expectancy, 2017")
```

## 

```{r, echo=FALSE}
delay_2019 %>% 
  filter(min_delay>0) %>% 
  ggplot(aes(min_delay)) + 
  geom_histogram(fill = "darkturquoise", color = "navy", binwidth = 10) +
  #scale_x_log10() + 
  theme_bw(base_size = 14) + 
  xlab("delay (minutes)") + 
  labs(title = "Delay times, TTC subway 2019")
```

## Making the histogram more informative

```{r, echo=FALSE}
ggplot(data = delay_2019) + 
  geom_histogram(aes(x = min_delay, y = ..density.., fill = line), position = 'dodge', bins = 10) + scale_x_log10()+
  theme_bw(base_size = 14) + 
  xlab("delay (minutes)") + 
  labs(title = "Delay times, TTC subway 2019",
       subtitle = "by line")
```

## Bar charts

Shows summary measures across values of a **categorical** (qualitative) variable

- Illustrate the value of a particular outcome in a particular category
- The 'value' can be counts, but could also be a summary measure (e.g. mean)
- The value is again shown by a rectangle of height equal to the value
- Bar carts can be plotted vertically or horizontally
- In the vertical setting, the x axis is the categories and the y axis is the value of the quantitative variable

##

```{r, echo = FALSE}
delay_2019 %>% 
  filter(min_delay>0) %>% 
  group_by(line) %>% 
  count() %>% 
  mutate(line = factor(line, levels = c("YU", "BD", "SRT", "SHP"),
                       labels = c("YU", "BD", "SRT", "SHP"))) %>% 
  ggplot(aes(x = line, y = n, fill = line)) + 
  geom_col() + 
  theme_bw(base_size = 14)+
  theme(legend.position = 'none') +
  ylab("count")+
  ggtitle("Number of delays by line, 2019")
```

## Same but horizontal

```{r, echo = FALSE}
delay_2019 %>% 
  filter(min_delay>0) %>% 
  group_by(line) %>% 
  count() %>% 
  mutate(line = factor(line, levels = c("YU", "BD", "SRT", "SHP"),
                       labels = c("YU", "BD", "SRT", "SHP"))) %>% 
  ggplot(aes(x = line, y = n, fill = line)) + 
  geom_col() + 
  theme_bw(base_size = 14)+
  theme(legend.position = 'none') +
  ylab("count")+
  ggtitle("Number of delays by line, 2019")+
  coord_flip()
```

## Showing mean delay time

```{r, echo = FALSE}
delay_2019 %>% 
  filter(min_delay>0) %>% 
  group_by(line) %>% 
 summarise(n = mean(min_delay)) %>% 
  mutate(line = fct_reorder(as.factor(line), n)) %>% 
  ggplot(aes(x = line, y = n, fill = line)) + 
  geom_col() + 
  theme_bw(base_size = 14)+
  theme(legend.position = 'none') +
  ylab("mean delay (mins)")+
  ggtitle("Mean length of delay by line, 2019") + 
  coord_flip()
```

## More complicated example

```{r, echo = FALSE}
delay_2019 %>% 
  filter(min_delay>0) %>% 
  group_by(line, station) %>% 
  summarise(mean_delay = median(min_delay), n_obs = n()) %>% 
  filter(n_obs>1) %>% 
  arrange(line, -mean_delay) %>% 
  slice(1:5) %>% 
  ggplot(aes(station, mean_delay)) + geom_col() + coord_flip() + facet_wrap(~line, scales = "free_y")+
  ggtitle("Top 5 station by median delay") + 
  xlab("median delay (mins)")+
  theme_bw(base_size = 14)
```

<!-- ## A lego example -->

<!-- \tiny -->
<!-- ```{r} -->
<!-- lego <- tibble(color = c("green", "white", "pink",  -->
<!--                          "yellow", "blue", "light green", "orange"), -->
<!--                count = c(6,5,4,3,2,2,1)) -->
<!-- ``` -->

<!-- ```{r, echo = FALSE} -->
<!-- lego <- lego %>%  -->
<!--   mutate(color = fct_reorder(color, -count)) -->
<!-- ``` -->

<!-- \centering -->
<!-- \includegraphics[width = 0.8\textwidth]{../fig/lego_bar.jpeg} -->

<!-- ## -->
<!-- \tiny -->
<!-- ```{r, echo = TRUE, fig.height = 4, fig.width = 6} -->
<!-- ggplot(lego, aes(color, count, fill = color)) +  -->
<!--   geom_bar(stat="identity") +  -->
<!--   scale_fill_manual(values = c("#70961c", "white",  -->
<!--                                "#ee5e4f", "#d5c47c", "#008db3", "#a5d395", "#d35800")) +  -->
<!--   theme(legend.position = 'none', -->
<!--         panel.background = element_rect(fill = "#d7d3c9", -->
<!--                                 colour = "#d7d3c9", -->
<!--                                 size = 0.5, linetype = "solid")) + -->
<!--   ylab("number of blocks")+xlab("") -->
<!-- ``` -->


## Box plots

Good for showing summaries of **quantitative** variables across different **categorical** groups. 


- Visualizing quartiles (25/50/75 percentiles) of quantitative data
- Boxes show the IQR and median
- Whiskers show values outside IQR (in R/`ggplot`, default is 1.5*IQR)
- Outliers may be shown with individual dots
- In the vertical case, the x axis is the categories and the y axis is the quantitative variable

## 

```{r, echo = FALSE}
country_ind %>% 
  mutate(region = fct_reorder(region, -life_expectancy)) %>% 
  ggplot(aes(x = region, y = life_expectancy, fill = region)) + 
  geom_boxplot() +
  scale_fill_viridis_d()+
  theme_bw(base_size = 14)+
  theme(legend.position = 'none', axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  xlab("")+
  ylab("")+
  ggtitle("Life expectancy (years) by region of the world")
```

## Could also do horizontal


```{r, echo = FALSE}
country_ind %>% 
  mutate(region = fct_reorder(region, -life_expectancy)) %>% 
  ggplot(aes(x = region, y = life_expectancy, fill = region)) + 
  geom_boxplot() +
  scale_fill_viridis_d()+
  theme_bw(base_size = 14)+
  theme(legend.position = 'none')+
  xlab("")+ylab("")+
  coord_flip()+
  ggtitle("Life expectancy (years) by region of the world")
```





## Line plots

Best used to describe values of a **quantitative** variable (on y axis) across sequential values of another **quantitative** variable on the x axis

- Plots a series of values of a quantitative variable connected together by a line
- Useful to visualize trends over time

##

```{r, echo = FALSE}
country_ind %>%
  filter(country %in% c("Canada", "United States of America")) %>% 
  ggplot(aes(year, tfr, color = country)) + 
  geom_line(lwd = 1.2) + 
  geom_point(size = 2) + 
  ggtitle("Total fertility rate, Canada and the US") + 
  scale_x_continuous(breaks = 2009:2017) + theme_bw(base_size = 14)
```


## Scatter plots

Shows relationship between two different **quantitative** variables

- Uses dots to represent values for two different **quantitative** values
- The position of each dot on the x and y axis indicates values for an individual data point
- Extremely useful in visualizing the relationship between two quantitative variables

##

```{r, echo = FALSE}
country_ind %>% 
  filter(year == 2017) %>% 
  ggplot(aes(tfr, life_expectancy)) + 
  geom_point() + 
  ggtitle("TFR versus life expectancy, 2017")+
  theme_bw(base_size = 14)+ 
  ylab("life expectancy (years)") + 
  xlab("TFR (births per woman)")
```

##

```{r, echo = FALSE}
country_ind %>% 
  filter(year == 2017) %>% 
  ggplot(aes(tfr, life_expectancy, color = region)) + 
  geom_point() + 
  ggtitle("TFR versus life expectancy, 2017")+
  theme_bw(base_size = 14)+ 
  ylab("life expectancy (years)") + 
  xlab("TFR (births per woman)") + 
  scale_color_viridis_d()
```

## Aside: another summary measure

Based on the previous graphs, evidence to suggest a relationship between TFR and life expectancy

- as TFR goes up, life expectancy tends to go down
- life expectancy is **negatively correlated** with TFR

**Correlation** is the statistical measure of the relationship between two variables. **Pearson's correlation coefficient**, $r_{xy}$ summarizes this relationship into one number. For an observation sample of two random variables $x_1, x_2, \dots, x_n$ and $y_1, y_2, \dots, y_n$, 

$$
r_{xy} = \frac{\sum^n_{i=1}(x_i - \bar x)(y_i - \bar y)}{\sqrt{\sum^n_{i=1}(x_i - \bar x)^2}\sqrt{\sum^n_{i=1}(y_i - \bar y)^2}}
$$

## Correlation coefficients in R

Easiest to use the function `cor`

\footnotesize
```{r}
country_ind_2017 <- country_ind %>% filter(year==2017)
country_ind_2017 %>% 
  select(tfr, life_expectancy) %>% 
  summarize(correlation = cor(tfr, life_expectancy))

# alternative code
# cor(country_ind_2017$tfr, country_ind_2017$life_expectancy) 
```

##

```{r, echo = FALSE}
country_ind %>% 
  filter(year == 2017) %>% 
  ggplot(aes(tfr, life_expectancy)) + 
  geom_point() + 
  ggtitle("TFR versus life expectancy, 2017")+
  theme_bw(base_size = 14)+ 
  ylab("life expectancy (years)") + 
  xlab("TFR (births per woman)")+
  geom_smooth(method = "lm")
```

## Plot your data!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!

The correlation coefficient tells us in a single number that there is a negative relationship observed between TFR and life expectancy. So why bother plotting at all?

## Anscombe's quartet

```{r, echo = FALSE, fig.width=6, fig.height=3}
datasets::anscombe %>% 
  select(x1:x4) %>% 
  pivot_longer(x1:x4) %>% 
  mutate(name = as.numeric(str_remove(name, "x"))) %>% 
  rename(dataset = name,
         x = value) %>% 
  group_by(dataset) %>% 
  mutate(i = 1:n()) %>% 
  left_join(datasets::anscombe %>% 
  select(y1:y4) %>% 
  pivot_longer(y1:y4) %>% 
  mutate(name = as.numeric(str_remove(name, "y"))) %>% 
  rename(dataset = name,
         y = value) %>% 
  group_by(dataset) %>% 
  mutate(i = 1:n())) %>% 
  ggplot(aes(x, y)) + geom_point() + facet_wrap(~dataset) + 
  geom_abline(intercept = 3, slope = 0.5, color = "royalblue", lwd = 1.2)+
  theme_bw(base_size = 14)
```

## 

\includegraphics{../fig/dinos.png}
\tiny
Source: https://www.autodeskresearch.com/publications/samestats

## Where are the pie charts?

Don't use pie charts!, Humans are inherently bad at judging angles, which is what you have to do with a pie chart. Use a bar chart instead. 

```{r, echo = FALSE}
p1 <- delay_2019 %>% 
  group_by(line) %>% 
  tally() %>% 
  mutate(prop = n/sum(n)) %>% 
  mutate(ypos = cumsum(prop) - 0.5*prop) %>% 
  #mutate(line = fct_reorder(line, prop)) %>% 
  ggplot(aes(x = "", y = prop, fill = line)) + geom_bar(stat = "identity", width = 1) +
  coord_polar("y")+
  theme_void()
p2 <- delay_2019 %>% 
  group_by(line) %>% 
  tally() %>% 
  mutate(prop = n/sum(n)) %>% 
  mutate(ypos = cumsum(prop) - 0.5*prop) %>% 
  mutate(line = fct_reorder(line, -prop)) %>% 
  ggplot(aes(x = line, y = prop, fill = line)) + geom_bar(stat = 'identity')+
  theme_bw()

library(patchwork)
p1+p2
```


# Introduction to `ggplot`

## `ggplot`

- `ggplot` is the graphing package that goes with the `tidyverse` in R
- Very powerful to make a wide range of graphics
- Every graph so far this lecture was done in `ggplot`
- `ggplot` code works in layers, with each layer adding complexity
    + start with defining dataset and different variables
    + add on type of plot
    + scales
    + layout (facets)
    + themes, fonts, sizes...

More practice in lab, but here's a starting example

## Reproducing the TFR verus life expectancy chart, colored by region

```{r, echo = FALSE}
country_ind %>% 
  filter(year == 2017) %>% 
  ggplot(aes(tfr, life_expectancy, color = region)) + 
  geom_point() + 
  ggtitle("TFR versus life expectancy, 2017")+
  theme_bw(base_size = 14)+ 
  ylab("life expectancy (years)") + 
  xlab("TFR (births per woman)") + 
  scale_color_viridis_d()
```

## Data

\tiny

```{r}
# read in the data
country_ind <- read_csv("../../data/country_indicators.csv")
country_ind
```
```{r}
# filter to just be 2017
country_ind_2017 <- country_ind %>% filter(year==2017)
```


## A blank canvas
`aes` stands for aesthetic and tells ggplot the main characteristics of your plot (x, y, and if the color or fill vary by group)

\tiny
```{r, fig.width=4, fig.height = 2.2}
plot1 <- ggplot(data = country_ind_2017, aes(x = tfr, y = life_expectancy))

#print
plot1
```

## Add the points
Add layers with ggplot using the `+`
\tiny
```{r, fig.width=4, fig.height = 2.2}
plot1 <- ggplot(data = country_ind_2017, aes(x = tfr, y = life_expectancy)) + 
  geom_point()

plot1
```

## Tidy up labels

\tiny
```{r, fig.width=4, fig.height = 2.2}
plot1 <- ggplot(data = country_ind_2017, aes(x = tfr, y = life_expectancy)) + 
  geom_point()+
  xlab("TFR (births per woman)")+
  ylab("life expectancy (years)")
  
plot1
```

## Title

\tiny
```{r, fig.width=4, fig.height = 2.2}
plot1 <- ggplot(data = country_ind_2017, aes(x = tfr, y = life_expectancy)) + 
  geom_point()+
  xlab("TFR (births per woman)")+
  ylab("life expectancy (years)")+
  ggtitle("TFR versus life expectancy, 2017")
  
plot1
```

## Change color of points
to see all colors, type `colors()`

\tiny
```{r, fig.width=4, fig.height = 2.2}
plot1 <- ggplot(data = country_ind_2017, aes(x = tfr, y = life_expectancy)) + 
  geom_point(color = "red")+
  xlab("TFR (births per woman)")+
  ylab("life expectancy (years)")+
  ggtitle("TFR versus life expectancy, 2017")
  
plot1
```


## Coloring by group 

This goes in the `aes()` because it **depends on the data**
\tiny
```{r, eval = FALSE}
plot1 <- ggplot(data = country_ind_2017, aes(x = tfr, y = life_expectancy, color = region)) + 
  geom_point()+
  xlab("TFR (births per woman)")+
  ylab("life expectancy (years)")+
  ggtitle("TFR versus life expectancy, 2017")
  
plot1
```

## 
\tiny
```{r, echo = FALSE}
plot1 <- ggplot(data = country_ind_2017, aes(x = tfr, y = life_expectancy, color = region)) + 
  geom_point()+
  xlab("TFR (births per woman)")+
  ylab("life expectancy (years)")+
  ggtitle("TFR versus life expectancy, 2017")
  
plot1
```


## Change theme (optional) and size of points
\tiny
```{r, eval = FALSE}
plot1 <- ggplot(data = country_ind_2017, aes(x = tfr, y = life_expectancy, color = region)) + 
  geom_point(size =2)+
  xlab("TFR (births per woman)")+
  ylab("life expectancy (years)")+
  ggtitle("TFR versus life expectancy, 2017")+
  theme_bw(base_size = 14)
  
plot1
```

## 
\tiny
```{r, echo = FALSE}
plot1 <- ggplot(data = country_ind_2017, aes(x = tfr, y = life_expectancy, color = region)) + 
  geom_point(size =2)+
  xlab("TFR (births per woman)")+
  ylab("life expectancy (years)")+
  ggtitle("TFR versus life expectancy, 2017")+
  theme_bw(base_size = 14)
  
plot1
```

## Change color scheme
`viridis` and `brewer` both good options
\tiny
```{r, eval = FALSE}
plot1 <- ggplot(data = country_ind_2017, aes(x = tfr, y = life_expectancy, color = region)) + 
  geom_point(size =2)+
  xlab("TFR (births per woman)")+
  ylab("life expectancy (years)")+
  ggtitle("TFR versus life expectancy, 2017")+
  theme_bw(base_size = 14)+
  scale_color_viridis_d()
  
plot1
```

## 
\tiny
```{r, echo = FALSE}
plot1 <- ggplot(data = country_ind_2017, aes(x = tfr, y = life_expectancy, color = region)) + 
  geom_point(size =2)+
  xlab("TFR (births per woman)")+
  ylab("life expectancy (years)")+
  ggtitle("TFR versus life expectancy, 2017")+
  theme_bw(base_size = 14)+
  scale_color_viridis_d()
  
plot1
```


## Summary

- EDA and data visualization is often just as informative and important as statistical analysis
- It is essential to understand the structure of your data, missing-ness, any outliers/issues, and the raw patterns in your data before deciding on your statistical analysis
- Plot, plot, plot
- Practice, practice, practice

## Summary

Plots:

- Bar charts for categorical/qualitative variables 
- Histograms, boxplots for one quantitative variable (potentially across multiple categories)
- Line plots and scatter plots for two quantitative variables (line plot when one is sequential)

## Lab


- Quick overview of the end of last week's lab (see video)
- Practice with `ggplot` and how to graph important types of plots in R.


## Data ideas

- IPUMS: https://ipums.org/
- ICPSR: https://www.icpsr.umich.edu/web/pages/ICPSR/thematic-collections.html
- CHASS SDA: https://datacentre.chass.utoronto.ca/
- Toronto Open Data Portal: https://open.toronto.ca/ or use `opendatatoronto` R package (ask for code)
- UN WPP: https://population.un.org/wpp/
- NBER: https://www.nber.org/research/data?page=1&perPage=50
