---
title: "SOC6707: Intermediate Data Analysis"
author: "Monica Alexander"
date: "Week 1: Introduction"
output: 
  beamer_presentation:
    slide_level: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
```

## Overview of today

- Overview of course
- Why learn statistics?
- Review concepts
- Introduction to R

# Overview of course

## Instructor and TA

Instructor: Monica Alexander (she/her)

- Email: monica.alexander@utoronto.ca
- Office hours: TBA

TA: Julia Igenfeld (she/her)

- Email: julia.ingenfeld@mail.utoronto.ca
- Office hours: TBA

## A bit about me

Me:

- statistics $\cap$ chemistry  $\rightarrow$ social science $\cap$ statistics
- 50/50 Statistical Sciences and Sociology departments
- Not Canadian (Australia $\rightarrow$ USA $\rightarrow$ Canada )


What I work on: a mix of demography, applied stats, epidemiology and computational social science


## Mode of delivery

For the most part, this course will be delivered in **online synchronous** format. 

- Lectures, tutorials, and office hours online through BBCollaborate (accessed through Quercus)
- Lectures will be recorded such that they can be accessed at a later date
- TBD: maybe some short pre-recorded materials

## Objectives

This course introduces statistical techniques and methods to analyze data to draw inferences about social processes. You will learn

- How to read in, describe, plot and analyze data in a statistical software that uses a programming language
- Some important methods of statistical analysis to explore relationships between social phenomena
- How to assess and evaluate the suitability and performance of statistical methods in different contexts

Methods covered include generalized linear models, Bayesian inference and multilevel models. 

## General philosophies

- Understanding your data
- Reproducibility
- Knowing when and when not to use methods 

This course will be very hands-on with coding and data munging. The learning curve for R is steep but will (hopefully?!) pay off. 

## Textbooks

There is no required textbook for this class. Some resources that might be useful:

- R for Data Science: https://r4ds.had.co.nz/ (free)
- Gelman, Andrew; Hill, Jennifer, and Vehtari, Aki. 2020. ‘Regression and Other Stories’ (This is around $70 on Book Depository).

## Software

We will be using the programming language R in this course, through RStudio. 

- Free to download and install
- More info on how to install these on Quercus
- More introduction later on in lecture

## Assessment

Overview

- Three assignments (3x15% = 45%)
- Mid-term (20%)
- Research project (35%)

## Assignments

- Data analysis with R
- Interpretation
- Hand in code, instructor/TA should be able to run without errors


## Mid-term exam

- To be taken online through Quercus
- Multiple choice and short answer questions
- Will need to be able to interpret R output

## Research Project

Choose a data set, research question, and analysis approach

In the research project you will

- Develop a research question based on data set of choice
- Analyze data using methods learned in class
- Present, interpret and summarize findings

## Research Project

Worth a total of 35%, but will be graded in four parts:

1. Research question, variables to be used (5%) due with A1
2. EDA (5%) due with A2
3. Analysis (5%) due with A3
4. Final report, which incorporates 1-3 (20%), due at end of semester. 

More detail in course outline, and as we go along.

## Lecture + Lab

- Each week I will lecture for about 1-1.5 hours, then we will have a lab with hands-on practice in R. 

## Course Policies

- **Communication**: First, see if you can answer your question by checking the syllabus. Second, try to ask questions during class, tutorials, or office hours. Third, there will be a discussion board on Quercus. Fourth, email myself or your TA (please include the course number in the subject line)
- **Accessibility**: visit http://studentlife.utoronto.ca/accessibility as soon as possible.

## We're all out here doing our best

- The current situation makes both learning and teaching challenging
- Try to be understanding of everyone's sub-optimal situation
- Communication is key
- There may be guest appearances from my toddler

## Doing okay?

Whether it is the chaos of 2020 getting you down, or something else, it is always okay to reach out for support!

- My Student Support Program – My SSP – mental health support for all U of T students. Free, confidential, immediate support. Available 24/7 in multiple languages. Download the My SSP App or call 1-844-451-9700. uoft.me/myssp
- Call Good2Talk. Free, confidential helpline with professional counseling, information and referrals for mental health, addictions and well-being, 24/7/365 1-866-925-5454


# Why learn statistics?

## Why learn statistics?

As sociologists, we are trying to understand different aspects of society. 

Statistical techniques give us a means to investigate and test research questions and policy impacts across different areas of people's lives.

Example research questions could include

- How is population mobility changing in the era of Covid-19?
- How do people cope with financial hardship?
- How does paid maternity leave affect women's workforce participation?
- Does volunteering increase your sense of wellbeing?


## Why learn statistics

**It's not just learning what you could do with data, it's learning what not to do with data**

- How biases and selection can give misleading conclusions
- When is it inappropriate to use certain techniques

**It's not just to support your own arguments, it's learning how to assess other people's arguments**

- Statistics, data analysis and visualization is an art form
- Cutting through the lies, damned lies and (misused) statistics

## Misleading statistics I

\begin{figure}
\centering
\includegraphics[width = 0.7\textwidth]{../fig/truman.png}
\end{figure}

- Truman won the Presidential election in 1948
- This is a photo of Truman holding an up an erroneous headline
- Based on phone survey which predicted overwhelming win for Dewey
- What went wrong?

## Misleading statistics II

\begin{figure}
\centering
\includegraphics[width = 0.7\textwidth]{../fig/planes.png}
\end{figure}

- Abraham Wald in WWII
- Want to place armor on planes in most effective place
- Gathered data from planes returning from battle and observed bullet holes
- Most holes in the fuselage, not so many in the engines
- Where should armor go?


## Misleading statistics III

\begin{figure}
\centering
\includegraphics[width = 0.7\textwidth]{../fig/temp_1.png}
\end{figure}

## Alternative

\begin{figure}
\centering
\includegraphics[width = 0.8\textwidth]{../fig/temp_2.jpeg}
\end{figure}

# Review

## Populations

At the core of statistical methods is wanting to say something about a **population** of interest.

What is a population? Depends on the context of study

- Everyone enrolled in university in Canada
- Everyone at UofT
- Everyone studying graduate-level sociology at UofT
- Everyone in this class

## Samples

Say we want to study the relationship between hours studied and overall achievement for all university students in Canada. 

- Not really plausible to get data on this for the whole of Canada.
- In reality, we would collect data on a **subset** or **sample** of the population and try and generalize to the whole of Canada. 
- With statistics, we are going to make conclusions based on what we see in the sample that we hope will be true for the population.

## Samples

Our example: the relationship between hours studied and overall achievement for all university students in Canada.

- We could plausibly measure the hours studied and overall achievement in SOC6707
- This class would be a sample of the population of interest, because you are all university students in Canada.

Is it a good sample? What do I mean by good?

## Sampling techniques

Terminology:

- **Element**: An element is an object or case, the unit on which a measurement is made.
- **Population**: The population is a collection of elements about which we wish to make an inference
- **Sampling units**: The sampling units are non-overlapping collections of elements in the population.
- **Sampling frame**: The sampling frame is a list of the elements or, for more complex samples, a list
of the sampling units.
- **Sample**: A sample is a subset of the elements drawn from the population using one of several
sampling methods.

What are each of these in our example?

## Sampling techniques

Include: 

- **Simple Random Sampling (SRS)**: A random sample is sometimes defined as a sample in which all possible elements have an equal
chance of occurring.
- **Stratified Sampling**: based on variable of interest
- **Cluster Sampling**: SRS within clusters (e.g. districts within a province, schools within districts)
- **Convenience Sampling**

What is the sampling method in our example?


## Two main domains of statistics

- **Descriptive statistics**: uses the data to provide descriptions of the population, either through numerical calculations or graphs or tables. 
- **Inferential statistics**: makes inferences and predictions about a population based on a sample of data taken from the population in question.

We will cover both types in this course. Understanding patterns in descriptive statistics is essential to doing good inferential statistics

## Variables

Traits, characteristics, outcomes that we are interested in. e.g.

- hours of study
- course grade
- province of residence
- age
- self-reported health

## Variables

Often we are interested in studying the relationship between two or more variables.

- The **outcome** of interest is the **dependent variable** 
- Variables **used to explain the outcome** can be called 
    + independent variables
    + explanatory variables
    + covariates
    + predictors
    
I will use these terms interchangeably. 

What is the independent and dependent variable in our example?

## Types of measurement of variables

- **Quantitative**: has a numeric meaning
    + **Continuous**: any possible number
    + **Discrete**: possible values can assume only certain values, usually the counting numbers
- **Qualitative**: categorical, no numeric meaning

What are the types of variables in our example?

## Random variables

- A **random variable** is a variable whose values depend on the outcomes of a random process. 
- For our purposes, the "random process" is taking a random sample of a population
- For example, consider the variable annual income: 
    + We randomly select someone from the population and note their income.
    + The value of this depends on the person who was selected
    + If we randomly selected someone else again, it's likely that the income value would be different

RVs are the basis of probability and statistical inference! Much more on their properties later.

## Some common symbols and notation

We will see a fair bit of math notation in this class, but there are some common notation that will come up over and over. To start with

- Population size = $N$
- Sample size = $n$
- A particular individual in a sample denoted by index $i$
- Random variables (note these are captials!):
    + Dependent variable: $Y$
    + Independent variables $X$
- A set of random variables for individuals $i = 1, 2, \dots, n$: ${X_1, X_2, \dots, X_n}$
- Specific values or outcomes of the corresponding random variables:
    + Dependent variable: $y$
    + Independent variables $x$

# Summary measures of quantitative data

## Summary measures of quantitative data

Pretend you have a set of observations of a quantitative variable, e.g. everyone's height in this class. We often want to **summarize** our set of observations with one or more numbers. Often interested in:

- Measures of central tendency, i.e. what would we expect someone's height to be, what's the most common height
- Measures of spread, i.e. what are the ranges of heights observed, what is the deviation of heights away from the expected height?

## Measures of central tendency

- **Mean**: the average
    + Population mean usually denoted as $\mu$
    + Sample mean denoted with a bar e.g. $\bar{x}$

$$
\bar{x} = \frac{\sum_{i=1}^n x_i}{n}
$$

- **Median**: the value for which 50% of the sample is below and 50% of the sample is above. It is the 50% percentile. To calculate
    + Order set of values from smallest to largest
    + find the middle number
    
- **Mode**: the value that occurs the most frequently


## Example

\footnotesize
```{r}
x <- c(4,6,2,1,6,2,76,3,2,56,10,1,4,5,2,15,32)
sort(x)
```

\normalsize

- Mean = ?
- Median = ?
- Mode = ?

## Example

```{r}
mean(x)
median(x)
# the mode is a bit more tricky
table(x) 
names(sort(table(x), decreasing = TRUE)[1])
```


## Measures of variability

- **Range**: The difference between the minimum and maximum value
- **Interquartile range**: The difference between the 25% and 75% percentiles. To calculate
    + Order set of values from smallest to largest
    + Separate into quarters
    + Find the first quarter (Q1) and third quarter (Q3)
    + IQR = Q3 - Q1

## Measures of variability

- **Variance**: average of the squares of the deviations
    + Population variance: $\sigma^2$
    + Sample variance: $s^2$
    
$$
s^2 = \frac{\sum_{i=1}^n (x_i- \bar{x})^2}{n-1}
$$

- **Standard deviation**: average of the deviations 
    + Population standard deviation: $\sigma$
    + Sample standard deviation: $s$
    
$$
s = \sqrt{\frac{\sum_{i=1}^n (x_i- \bar{x})^2}{n-1}} = \sqrt {s^2}
$$

## Example R code

\tiny
```{r}
sort(x)
max(x) - min(x)
IQR(x)
var(x)
sd(x)
#check this is the same
sqrt(var(x))
```



# Introduction to R

## R and RStudio

- You will need to download and install both R and RStudio
- More info on Quercus
- Try and do this before the lab

## What is R?

- R is a programming language for statistical computing and graphics
- Using R is like speaking another language (but you type it)

You may have used other programs to do statistical calculations before (Excel, SPSS)

- With R you have to give the computer typed commands in order for it to do stats (rather than clicking buttons)
- Much more powerful methods

\begin{figure}
\centering
\includegraphics[width = 0.2\textwidth]{../fig/Rlogo.png}
\end{figure}

## What is RStudio?

- RStudio is an integrated development environment for R. 
- It makes it easier to write R code and visualize inputs and outputs. 
- You will need to download and install both R and RStudio, details in course outline (also in lab 1)

Think of a car analogy:

- R is the engine
- RStudio is the car dashboard (steering wheel, controls etc)

\begin{figure}
\centering
\includegraphics[width = 0.2\textwidth]{../fig/rstudio.png}
\end{figure}

## RStudio

\begin{figure}
\centering
\includegraphics[width = 1\textwidth]{../fig/rstudio_screen.png}
\end{figure}

## Writing code in R


1. R Console: Executes each line of code as you go; does not save code for later use
2. R Script: Saves code and comments in a file so you can select some or all of the code in a
script file to run; does not include output
3. R Markdown: A file which combines text and chunks of R code (which can be executed independently). This allows you to see output without "knitting" the whole file.

We will focus on number 3.

## R Markdown documents

\begin{figure}
\centering
\includegraphics[width = 1\textwidth]{../fig/example_rmd.png}
\end{figure}

## Knitted to a PDF

\begin{figure}
\centering
\includegraphics[width = 1\textwidth]{../fig/example_knitted.png}
\end{figure}

## R Packages

- A lot of people have written **R packages**, which are add ons to base R that increase the functionality

Think of a phone analogy:

- R/RStudio is a phone
- R packages are apps

We will be using a few different R packages quite a lot during the course, e.g.

- `dplyr` (data manipulation)
- `ggplot2` (graphing)

These and other packages can be downloaded through downloading the `tidyverse` package (you will do this in the lab)

## R code

Create a vector of numbers:

```{r, echo = TRUE}
x <- c(1,3,5,7,9)
```

Calculate summary statistics:

```{r, echo = TRUE}
mean(x)
max(x)
IQR(x)
```

## Reading in and manipulating real data

```{r, message= FALSE, echo = FALSE}
library(tidyverse)
gss <- read_csv(here("data/gss.csv"))
```

```{r, message= FALSE, echo = TRUE, eval = FALSE}
library(tidyverse)
gss <- read_csv(file = "data/gss.csv")
```

- `read_csv` is a **function** from the `tidyverse` package
- We are assigning the contents of the file to an object called `gss`
- The `gss` object is a data frame or **tibble** that contains all the GSS data
- We can now use other functions to manipulate and analyze the GSS data in R


## Selecting a column with `select()`

```{r}
select(gss, age)
```


## The pipe %>%

\footnotesize
```{r, echo = TRUE}
gss %>% 
  select(age)
```

- Read as "and then"
- So above we are taking the `gss` data **and then** selecting the age column

## More than one pipe / Important functions

Arrange

\footnotesize
```{r}
gss %>% 
  select(age) %>% 
  arrange(age)
```

## Important functions for manipulating data

\footnotesize

Arrange by descending order
```{r}
gss %>% 
  select(age) %>% 
  arrange(-age)
```


##  Important functions for manipulating data

Filter

\footnotesize
```{r}
gss %>% 
  select(age) %>% 
  filter(age<17)
```
##  Important functions for manipulating data

Mutate = add a new column

\footnotesize
```{r}
gss %>% 
  select(age) %>% 
  mutate(age_plus_1 = age+1)
```

##  Important functions for manipulating data

\footnotesize

```{r}
gss %>% 
  select(age) %>% 
  summarize(mean_age = mean(age))
```

## Important functions for manipulating data

- The pipe %>%
- `select` (columns)
- `filter` (rows)
- `arrange`
- `mutate`
- `summarize`

## Example: lego

\begin{figure}
\centering
\includegraphics[width = 0.45\textwidth]{../fig/lego_1.jpeg}
\end{figure}

\begin{figure}
\centering
\includegraphics[width = 0.45\textwidth]{../fig/lego_2.jpeg}
\end{figure}

## Make your own tibble

```{r}
lego <- tibble(color = c('brown', 'pink', 'red', 
                         'orange', 'yellow', 'light yellow',
                         'light green', 'green', 'blue',
                         'light blue', 'white'),
               number = c(1,1,6,3,7,4,1,7,4,2,7)
               )
```

## Make your own tibble

```{r}
lego
```

## Summary statistics

Find the mean number of blocks

```{r}
lego %>% 
  summarize(mean_blocks = mean(number))
```

Will leave it as an exercise to find median, mode, standard deviation, minimum and maximum. (Hint: for the mode, use `arrange`)


## Preview: graphing


\tiny
```{r, echo=TRUE, warning=FALSE, fig.height=6}
ggplot(data = gss %>% filter(marital_status!= "NA"), aes(age)) + 
  geom_histogram(position = "dodge", fill = "lightblue", bins = 20) + 
  facet_wrap(~marital_status) 
```

## Why you should learn R

- Free, open, reproducible, large community
- Used in industry
- Relatively easier to implement powerful stat methods
- Once you get the hang of it, can use to make
    + slides 
    + websites (e.g. [mine](https://www.monicaalexander.com/))
    + interactive applications (e.g. [here](https://monica-alexander.shinyapps.io/foster_care/))

## This week's lab

You should hopefully have R and RStudio already installed. If not, this is the first step!

- Learn how to make an RMarkdown file and execute code
- Install and load `tidyverse` package
- Read in GSS
- Practice important functions

## Where to get help

- Intro to R: 
    + R4DS is the most relevant textbook for learning "tidyverse" R
    + Telling stories with data: https://www.tellingstorieswithdata.com/01-03-r_essentials.html
- Lab time and office hours
