---
title: "SOC6707 Intermediate Data Analysis"
author: "Monica Alexander"
date: "Week 9: Introduction to hierarchical models"
output: 
  beamer_presentation:
    slide_level: 2
header-includes:
  - \usepackage[all,poly,ps,color]{xy}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, size = '\tiny')
```

## Hierarchical models

- Hierarchical models used to estimate parameters in settings where there is a hierarchy of nested populations. 
- Many problems have a natural hierarchy e.g.

    + patients within hospitals
    + school kids within classes within schools 
    + maternal deaths within countries within regions within the world
    

- Want to get estimates of underlying parameters of interest (e.g. probability of dying, test score, risk of disease) accounting for the hierarchy in the data
- A natural framework for including information at different levels of the hierarchy


## Radon example 

- Radon is a naturally occurring radioactive gas.
-  Its decay products are also radioactive; in high concentrations, they
can cause lung cancer (several 1000 deaths/year in the USA).
- Radon levels vary greatly across US homes.
- Data: radon measurements in over 80K houses throughout the US.
- Hierarchy: houses observed in counties.
- Potential predictors: floor (basement or 1st floor) in the house, soil uranium
level at country level.

```{r, include=FALSE}
library(tidyverse)
library(kableExtra)
d <- read.table(url("http://www.stat.columbia.edu/~gelman/arm/examples/radon/srrs2.dat"), header=T, sep=",")
```

## Radon dataset

Selected rows and columns

```{r}
head(d %>% select(idnum, state, county, basement ,activity)) %>% kable()
```


- 12,777 observations from `r length(unique(d$county))` counties

What might we want to estimate/predict?

## What might we want to estimate/predict?

- Expected radon level in a county
- Expected radon level in a county we did not have samples for
- Predicted radon level for a newly observed house in a particular county
- ...?

## Let's introduce some notation

- units $i = 1, \dots n$, the smallest items of measurement (household)
- outcome $y=\left(y_{1}, \ldots, y_{n}\right)$. The unit-level outcome being measure (log radon)
- groups $j = 1, \dots , J$ (counties)
- Indexing $j[i]$ (the county for house $i$)
- $x_i$ is an indicator, whether or not measurement was taken on basement (house level)
- $u_j$ is the uranium level in the soil (county level)

## Notation

Thinking about our usual regression set-up, we usual write as something like
$$
Y_i = \beta_0 + \beta_1X_i + \varepsilon_i
$$
Let's rewrite this as
$$
Y_i \sim N(\mu_i, \sigma^2)
$$
with 
$$
\mu_i = \beta_0 + \beta_1X_i
$$
and 
$$
\varepsilon_i \sim N(0, \sigma^2)
$$

These are equivalent. 

## A model for log radon

$$
Y_i \sim N(\mu_i, \sigma^2)
$$

- Note that $\mu_i = E(Y_i)$ i.e. the expected (log) radon level for a particular house $i$
- How to model $\mu_i$?
- Let's start simple (no covariates)
- Given we know house $i$ is in county $j$, how can we model $\mu_i$?

## One option: no pooling

Estimate the county-level mean for each county, using only the
data from that county. The model is 

$$y_{i} {\sim} N\left(\alpha_{j[i]}^{\text{no pool}}, \sigma_{y}^{2}\right)$$

- The "no pool" refers to treating each county separately, i.e. no pooling of information across counties
- The most appropriate estimator for this is the county mean, i.e. $\bar{y}_{j}$
- I.e. the expected level of log radon for a particular house $i$ in county $j$ is just the mean radon level for the county

## No pooling


```{r}
ybars <- d %>% 
  group_by(state, county) %>% 
  mutate(activity = ifelse(activity==0, 0.1, activity)) %>% 
  summarise(ybar  = mean(log(activity)),
            n = n()) %>% 
    group_by(n) 

d %>% 
  select(county, state, activity) %>% 
  right_join(ybars) %>% 
  filter(state=="PA") %>% 
  mutate(id = 1:n()) %>% 
  mutate(log_activity = log(activity)) %>% 
  filter(n<100, activity>0) %>% 
  ggplot(aes(n, log(activity))) + 
  geom_point(aes(color = "obs")) + 
  geom_point(aes(n, ybar, color = "county_mean"))  +
  scale_color_manual(name = "", values = c("obs" = "grey", "county_mean" = "red"))+
  ggtitle("radon measurments v sample size, Pennsylvania") + 
  theme_bw(base_size = 16) +
  ylab("log(radon activity)") + xlab("sample size of households in county")
```

What do you notice about this graph?

## Another option: complete pooling

- Maybe we believe that the expected radon level for a particular house is not going to vary by county
- Use the state mean as the best estimate for the means in each county.

Model is $$y_{i} \sim N\left(\mu, \sigma_{y}^{2}\right)$$

- i.e. expected log radon level is constant across state
- Best estimator here would just be state mean
- this is referred to "complete pooling" because information across all counties is pooled together 


## Complete pooling

```{r}
d %>% 
  select(county, state, activity) %>% 
  right_join(ybars) %>% 
  filter(state=="PA") %>% 
  mutate(id = 1:n()) %>% 
  mutate(activity = ifelse(activity==0, 0.1, activity)) %>% 
  mutate(log_activity = log(activity)) %>% 
  mutate(mean_act = mean(log(activity))) %>% 
  filter(n<100, activity>0) %>% 
  ggplot(aes(n, log(activity))) + geom_point(aes(color = "obs")) + 
  geom_point(aes(n, ybar, color = "county_mean"))  +
  geom_hline(aes(yintercept = mean_act, color = "state_mean")) + 
  scale_color_manual(name = "", values = c("obs" = "grey", "county_mean" = "red", "state_mean" = "black"))+
  ggtitle("radon measurments v sample size, Pennsylvania") + 
  theme_bw(base_size = 16) +
  ylab("log(radon activity)") + xlab("sample size of households in county")
```

Pros? Cons?

## A happy medium

\centering
\includegraphics[width = 0.5\textwidth]{../fig/goldilocks.jpeg}

- Ideally we want to allow expected county radon levels to differ
- But we also want to account for information across all counties and not treat counties as separate
- A solution: partial pooling via hierarchical modeling 

## Another option: hierarchical model

- The expected radon level in a particular house $i$ is 
- county means $\alpha_j$ come from some common distribution across a state
- there are some underlying parameters governing the distribution of $\alpha$'s, which are generally unknown
- middle ground between first two options, $\alpha$'s are similar but not the same

## Hierarchical model

The model is
\begin{eqnarray*}
y_{i}  & \sim & N\left(\alpha_{j[i]}, \sigma_{y}^{2}\right) \\ 
\alpha_{j} & \sim  & N\left(\mu_{\alpha}, \sigma_{\alpha}^{2}\right)
\end{eqnarray*}

- The $alpha_j$'s are themselves assumed to be from a common distribution
- $\mu_{\alpha}$ and $\sigma_{\alpha}$ are called **hyperparameters**

## Hierarchical model

\begin{xy}
          \xymatrix{
            & \mu_{\alpha}, \sigma_{\alpha} \ar[dl] \ar[d] \ar[drr] & & &  \\
            \alpha_1\ar[d] & \alpha_2\ar[d] &
            \cdots & \alpha_n\ar[d] \\
            y_{1}, y_2 & y_{3}, \dots y_{8} & & y_{12770}, \dots y_{12777} }
\end{xy}

Because of the hierarchical set-up, the resulting estimates for the
county means are in-between the no-pooling and complete-pooling
estimates.

## Compare to

- No pooling

\begin{xy}
          \xymatrix{ \alpha_1\ar[d] & \alpha_2\ar[d] &
            \cdots & \alpha_n\ar[d] \\
            y_{1}, y_2 & y_{3}, \dots y_{8} & & y_{12770}, \dots y_{12777} }
\end{xy}
        
- Complete pooling

\centering
\begin{xy}
          \xymatrix{
            & \mu \ar[dl] \ar[d] \ar[drr] & & &  \\
            y_{1} & y_{2} & \cdots & y_{12777}}        
  \end{xy}


## Many names

- Also known as multilevel models, I will probably flip between the two
- Fixed and random effects
    + $\alpha_j$'s commonly referred to as random effects, because they are modeled as random variables
    + fixed effects are parameters that don't vary by group, or to parameters that vary but are not modeled themselves (e.g. county/state indicator variables)
    
- random effects models, (generalized) linear mixed models, mixed effects models: often used as synonyms for multilevel models

## The effect of partial pooling in the radon case

```{r}
library(lme4)
d <- d %>% 
  mutate(activity = ifelse(activity==0, 0.1, activity),
         log_activity = log(activity))
mod <- lmer (log_activity ~ (1 | county), data = d %>% filter(state=="PA"))
part_pooled <- tibble(pp = coef(mod)$county[, '(Intercept)'])

ybars %>% 
  filter(state=="PA") %>% 
  bind_cols(part_pooled) %>% 
  ungroup() %>% 
  mutate(mean_act = mean(ybar) ) %>% 
  filter(n < 100) %>% 
  ggplot(aes(n, ybar)) + 
  geom_point(aes(color = "unpooled_mean"), size = 5) + 
  geom_point(aes(n, pp, color = "partially_pooled_mean"), size = 5) + 
  geom_hline(aes(yintercept = mean_act, color = "pooled_mean")) + 
  scale_color_manual(name = "", values = c("unpooled_mean" = "red", "partially_pooled_mean" = "blue", "pooled_mean" = "black")) + 
  theme_bw(base_size = 16) + 
  scale_x_log10() + 
  ylab("mean log radon") + xlab("sample size")
```

## The effect of partial pooling in the radon case

```{r}

ybars %>% 
  filter(state=="PA") %>% 
  bind_cols(part_pooled) %>% 
  ungroup() %>% 
  mutate(mean_act = mean(ybar) ) %>% 
  filter(n < 100) %>% 
  ggplot(aes(ybar, pp)) + 
  geom_point(aes(size = n)) +
  theme_bw(base_size = 16) + 
  geom_abline(intercept = 0, slope = 1) + ylab("partially pooled mean") + xlab("unpooled mean")
```

## The effect of partial pooling in the radon case

```{r}

ybars %>% 
  filter(state=="PA") %>% 
  bind_cols(part_pooled) %>% 
  ungroup() %>% 
  mutate(mean_act = mean(ybar) ) %>% 
  filter(n < 100) %>% 
  ggplot(aes(n, pp-ybar)) + 
  geom_point(size = 5) +
  theme_bw(base_size = 16) + 
  geom_hline(yintercept = 0) + ylab("difference") + xlab("sample size") + ggtitle("Difference in partially pool and unpooled means \nversus sample size")
```

## Where are we at

- Hierarchical models allow for 'information exchange' across groups 
- Has the effect 'shrinking' group means to the overall mean
- Shrinking effect is larger when the sample size in a particular group is smaller

## Why does this happen?

- It turns out that the estimate of the hierarchical mean $\hat{\alpha}_j$ is a weighted mean between information from that group $j$ and all the other groups:

$$
\hat{\alpha}_j = \frac{\frac{n_{j}}{\sigma_{y}^{2}} \bar{y}_{j}+\frac{1}{\sigma_{\alpha}^{2}} \mu_{\alpha}}{\frac{n_{j}}{\sigma_{y}^{2}}+\frac{1}{\sigma_{\alpha}^{2}}}
$$

# Adding covariates

## Adding covariates

For the radon example:

- The measurements are not exactly comparable across houses because
in some houses, measurements are taken in the basement, while in
other houses, 1st floor measurement are taken. (This is $x_i$)
- Additionally, county-level uranium measurements are probably
informative for across-county differences in mean levels. (This is $u_j$)

When adding covariates, need to think about 

- what level the covariate relates to 
- whether or not to model the effect hierarchically

## Including covariates at the unit level

- Let $x_i$ be the house-level first-floor indicator (with $x_i= 0$ for
basements, 1 otherwise).
- This is a house-level covariate
- We can include house-level predictors in the house-level mean as
follows:

$$
\begin{aligned}
y_{i} & \sim N\left(\alpha_{j[i]}+\beta x_{i}, \sigma_{y}^{2}\right), \text { for } i=1,2, \ldots, n \\
\alpha_{j} & \sim N\left(\mu_{\alpha}, \sigma_{\alpha}^{2}\right), \text { for } j=1,2, \ldots, J
\end{aligned}
$$

Note: we have varying intercepts but a constant slope

## Covariates at unit level

$$
\begin{aligned}
y_{i} & \sim N\left(\alpha_{j[i]}+\beta x_{i}, \sigma_{y}^{2}\right), \text { for } i=1,2, \ldots, n \\
\alpha_{j} & \sim N\left(\mu_{\alpha}, \sigma_{\alpha}^{2}\right), \text { for } j=1,2, \ldots, J
\end{aligned}
$$

- Estimate of $\beta$ is -0.693
- Estimate of $\mu_{\alpha}$ is 1.462

## County-specific intercepts

```{r, eval = TRUE}
d_pa <- d %>% filter(state=="MN")
n_counties <- length(unique(d_pa$county))
mod_hier <- lmer(log_activity ~ (1 | county) + floor, data = d_pa)
res <- coef(mod_hier)[[1]]["(Intercept)"]
ses <- attr(ranef(mod_hier)[[1]], "postVar")[,,1:n_counties]
df <- res %>% bind_cols(ses) 
colnames(df) <- c("alpha", "se")
df$county <- rownames(df)

df %>% 
  mutate(county = str_trim(county)) %>% 
  mutate(county = factor(county)) %>% 
  mutate(county = fct_reorder(county, alpha)) %>% 
  filter(county!="") %>% 
  ggplot(aes(county,alpha)) + geom_point() + coord_flip()+
  geom_errorbar(aes(ymin = alpha - se, ymax = alpha + se))
```

## Including covariates at the group level

- County-level log-uranium measurements $u_j$ are probably informative
for across-county differences in mean levels.
- We can include group-level predictors in the group-level mean as
follows:

$$
\begin{aligned}
y_{i}  & \sim N\left(\alpha_{j[i]}+\beta x_{i}, \sigma_{y}^{2}\right), \text { for } i=1,2, \ldots, n \\
\alpha_{j} & \sim N\left(\gamma_{0}+\gamma_{1} u_{j}, \sigma_{\alpha}^{2}\right), \text { for } j=1,2, \ldots, J
\end{aligned}
$$

## Adding covariates at group level

$$
\begin{aligned}
y_{i}  & \sim N\left(\alpha_{j[i]}+\beta x_{i}, \sigma_{y}^{2}\right), \text { for } i=1,2, \ldots, n \\
\alpha_{j} & \sim N\left(\gamma_{0}+\gamma_{1} u_{j}, \sigma_{\alpha}^{2}\right), \text { for } j=1,2, \ldots, J
\end{aligned}
$$

```{r}

cty <- read.table(url("http://www.stat.columbia.edu/~gelman/arm/examples/radon/cty.dat"), header = T, sep = ",")
cty <- cty %>% mutate(fips = 1000 * stfips + ctfips) %>% dplyr::select(fips, Uppm)

d_pa <- d %>% 
  filter(state=="MN") %>% 
  mutate(fips = stfips * 1000 + cntyfips) %>% 
  dplyr::select(fips, county, floor, activity) %>% 
  left_join(cty) %>% 
  mutate(log_uran = log(Uppm), log_activity = log(activity))


mod_hier_group <- lmer(log_activity ~ (1 | county) + floor + log_uran, data = d_pa)
#coef(summary(mod_hier_group))
#colMeans(coef(mod_hier_group)[[1]])
```

- Estimate of $\beta$ is -0.668
- Estimate of $\gamma_0$ is 1.407
- Estimate of $\gamma_1$ is 0.729


## County-specific intercepts

```{r}
res <- coef(mod_hier_group)[[1]]["(Intercept)"]
ses <- attr(ranef(mod_hier_group)[[1]], "postVar")[,,1:n_counties]
#se.coef(mod_hier_group)[[1]]["Intercept"]
df <- res %>% bind_cols(ses) 
colnames(df) <- c("alpha", "se")
df$county <- rownames(df)

df %>% 
  mutate(county = str_trim(county)) %>% 
  mutate(county = factor(county)) %>% 
  mutate(county = fct_reorder(county, alpha)) %>% 
  filter(county!="") %>% 
  ggplot(aes(county,alpha)) + geom_point() + coord_flip()+
  geom_errorbar(aes(ymin = alpha - se, ymax = alpha + se))
```

## County-level radon and uranium

```{r}

gamma0 <- coef(summary(mod_hier_group))[1,1]
gamma1 <- coef(summary(mod_hier_group))[3,1]

urans <- d_pa %>% 
  group_by(county, log_uran) %>% 
  slice(1) %>% 
  mutate(county = str_trim(county)) %>% 
  filter(county!="") %>% 
  dplyr::select(county, log_uran) 

df %>% 
  mutate(county = str_trim(county)) %>% 
  left_join(urans) %>% 
  filter(county!="") %>%
  mutate(alpha_hat =  log_uran*gamma1 + alpha) %>% 
  ggplot(aes(log_uran, alpha_hat)) + 
  geom_point() + 
  geom_errorbar(aes(ymin = alpha_hat - se, ymax = alpha_hat + se)) + 
  geom_abline(intercept = gamma0, slope = gamma1) +
  ylab("log radon level") + xlab("log uranium") + ggtitle("County radon levels versus Uranium, Minnesota")

  
```



# Varying slopes (TBD)


## What about letting the effect of $x_i$ vary by county?

- In last model, we assume that the difference between basement and first floor measurement is the same across
houses, no matter which county the house is in.
- What if that difference varies by county?


$$
\begin{aligned}
y_{i} & \sim N\left(\alpha_{j[i]}+\beta_{j[i]} x_{i}, \sigma_{y}^{2}\right), \text { for } i=1,2, \ldots, n \\
\alpha_{j} & \sim N\left(\mu_{\alpha}, \sigma_{\alpha}^{2}\right), \text { for } j=1,2, \ldots, J \\
\beta_{j} & \sim N\left(\mu_{\beta}, \sigma_{\beta}^{2}\right), \text { for } j=1,2, \ldots, J
\end{aligned}
$$

Allowing for varying slopes. 

## Allowing for varying slopes at unit level

$$
\begin{aligned}
y_{i} & \sim N\left(\alpha_{j[i]}+\beta_{j[i]} x_{i}, \sigma_{y}^{2}\right), \text { for } i=1,2, \ldots, n \\
\alpha_{j} & \sim N\left(\mu_{\alpha}, \sigma_{\alpha}^{2}\right), \text { for } j=1,2, \ldots, J \\
\beta_{j} & \sim N\left(\mu_{\beta}, \sigma_{\beta}^{2}\right), \text { for } j=1,2, \ldots, J
\end{aligned}
$$

- Estimate of $\mu_{\alpha}$ is 1.46
- Estimate of $\mu_{\beta}$ is -0.679

```{r}
mod_hier_slopes <- lmer(log_activity ~ floor + (1+floor|county), data = d_pa)
#coef(summary(mod_hier_slopes))
#coef(mod_hier_slopes)[[1]][1:10,]
#ranef(mod_hier_slopes)[[1]][1:10,]
#colMeans(coef(mod_hier_slopes)[[1]])
```

## County-specific intercepts

```{r}
res <- coef(mod_hier_slopes)[[1]]["(Intercept)"]
ses <- (attr(ranef(mod_hier_slopes)[[1]], "postVar"))[1,1,1:n_counties]
df <- res %>% bind_cols(ses) 
colnames(df) <- c("alpha", "se")
df$county <- rownames(df)

df %>% 
  mutate(county = str_trim(county)) %>% 
  mutate(county = factor(county)) %>% 
  mutate(county = fct_reorder(county, alpha)) %>% 
  filter(county!="") %>% 
  ggplot(aes(county,alpha)) + geom_point() + coord_flip()+
  geom_errorbar(aes(ymin = alpha - se, ymax = alpha + se))
```



## County-specific slopes

```{r}
res <- coef(mod_hier_slopes)[[1]]["floor"]
ses <- (attr(ranef(mod_hier_slopes)[[1]], "postVar"))[2,2,1:n_counties]
df <- res %>% bind_cols(ses) 
colnames(df) <- c("alpha", "se")
df$county <- rownames(df)

df %>% 
  mutate(county = str_trim(county)) %>% 
  mutate(county = factor(county)) %>% 
  mutate(county = fct_reorder(county, alpha)) %>% 
  filter(county!="") %>% 
  ggplot(aes(county,alpha)) + geom_point() + coord_flip()+
  geom_errorbar(aes(ymin = alpha - se, ymax = alpha + se))
```


# Hierarchical models in R

## Fitting hierarchical models in R

- Many different options and packages to do this
- Many powerful options fitting Bayesian hierarchical models using languages like Stan or JAGS (but no time!)
- We will be using the `lme4` package, which allows you to fit hierarchical models using commands that are a logical extension of `lm` and `glm`
- (So you will need to `install.packages(lme4)`)


## Radon levels in Minnesota

We will see in lab, but a brief introduction to notation.

What we would usually do:

```{r, eval = FALSE, echo = TRUE}
library(lme4)
d_mn <- d %>% filter(state=="MN")

mod_nopool <- lm(log_activity ~ county, data = d_mn)
mod_pool <- lm(log_activity ~ 1, data = d_mn)
```

Hierarchical model:

```{r, eval = FALSE, echo = TRUE}
mod_hier <- lmer(log_activity ~ (1 | county), data = d_mn)
```


## Radon levels in Pennsylvania

Adding covariates:

\tiny
```{r, eval = FALSE, echo = TRUE}
mod_hier <- lmer(log_activity ~ floor + log_uran + (1 | county), data = d_mn)
```




