---
title: "SOC6707 Intermediate Data Analysis"
author: "Monica Alexander"
date: "Week 10: Hierarchical models II"
output: 
  beamer_presentation:
    slide_level: 2
header-includes:
  - \usepackage[all,poly,ps,color]{xy}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, size = '\tiny')
```

## Recap

Last week we started talking about hierarchical models

- Account for hierarchical structure in data (e.g. houses within counties)
- 'Happy medium' between treating all groups separately versus all groups the same
- Group-level effects are treated as coming from a common distribution, which allows information to be pooled across groups
- Particularly useful when some groups have small sample sizes

## This week

- Varying slopes
- Hierarchical GLM (logistic regression)

## Last week

For radon, we got up to
$$
\begin{aligned}
y_{i}  & \sim N\left(\alpha_{j[i]}+\beta x_{i}, \sigma_{y}^{2}\right), \text { for } i=1,2, \ldots, n \\
\alpha_{j} & \sim N\left(\gamma_{0}+\gamma_{1} u_{j}, \sigma_{\alpha}^{2}\right), \text { for } j=1,2, \ldots, J
\end{aligned}
$$
```{r}
library(tidyverse)
library(kableExtra)
library(lme4)
d <- read.table(url("http://www.stat.columbia.edu/~gelman/arm/examples/radon/srrs2.dat"), header=T, sep=",")
d <- d %>% 
  mutate(activity = ifelse(activity==0, 0.1, activity),
         log_activity = log(activity))
cty <- read.table(url("http://www.stat.columbia.edu/~gelman/arm/examples/radon/cty.dat"), header = T, sep = ",")
cty <- cty %>% mutate(fips = 1000 * stfips + ctfips) %>% dplyr::select(fips, Uppm)

d_pa <- d %>% 
  filter(state=="MN") %>% 
  mutate(fips = stfips * 1000 + cntyfips) %>% 
  dplyr::select(fips, county, floor, activity) %>% 
  left_join(cty) %>% 
  mutate(log_uran = log(Uppm), log_activity = log(activity))
n_counties <- length(unique(d_pa$county))

```



# Varying slopes 


## What about letting the effect of $x_i$ vary by county?

- In last model, we assume that the difference between basement and first floor measurement is the same across
houses, no matter which county the house is in.
- What if that difference varies by county?


$$
\begin{aligned}
y_{i} & \sim N\left(\alpha_{j[i]}+\beta_{j[i]} x_{i}, \sigma_{y}^{2}\right), \text { for } i=1,2, \ldots, n \\
\alpha_{j} & \sim N\left(\mu_{\alpha}, \sigma_{\alpha}^{2}\right), \text { for } j=1,2, \ldots, J \\
\beta_{j} & \sim N\left(\mu_{\beta}, \sigma_{\beta}^{2}\right), \text { for } j=1,2, \ldots, J
\end{aligned}
$$

Allowing for varying slopes. 

## In R

\footnotesize
```{r, eval = FALSE, echo = TRUE}
mod_hier_slopes <- lmer(log_activity ~ floor + (1+floor|county), 
                        data = d_mn)
```


## Allowing for varying slopes at unit level

$$
\begin{aligned}
y_{i} & \sim N\left(\alpha_{j[i]}+\beta_{j[i]} x_{i}, \sigma_{y}^{2}\right), \text { for } i=1,2, \ldots, n \\
\alpha_{j} & \sim N\left(\mu_{\alpha}, \sigma_{\alpha}^{2}\right), \text { for } j=1,2, \ldots, J \\
\beta_{j} & \sim N\left(\mu_{\beta}, \sigma_{\beta}^{2}\right), \text { for } j=1,2, \ldots, J
\end{aligned}
$$

- Estimate of $\mu_{\alpha}$ is 1.46
- Estimate of $\mu_{\beta}$ is -0.679

```{r}
mod_hier_slopes <- lmer(log_activity ~ floor + (1+floor|county), data = d_pa)
#coef(summary(mod_hier_slopes))
#coef(mod_hier_slopes)[[1]][1:10,]
#ranef(mod_hier_slopes)[[1]][1:10,]
#colMeans(coef(mod_hier_slopes)[[1]])
```

## County-specific intercepts

```{r}
res <- coef(mod_hier_slopes)[[1]]["(Intercept)"]
ses <- (attr(ranef(mod_hier_slopes)[[1]], "postVar"))[1,1,1:n_counties]
df <- res %>% bind_cols(ses) 
colnames(df) <- c("alpha", "se")
df$county <- rownames(df)

df %>% 
  mutate(county = str_trim(county)) %>% 
  mutate(county = factor(county)) %>% 
  mutate(county = fct_reorder(county, alpha)) %>% 
  filter(county!="") %>% 
  ggplot(aes(county,alpha)) + geom_point() + coord_flip()+
  geom_errorbar(aes(ymin = alpha - se, ymax = alpha + se))
```



## County-specific slopes

```{r}
res <- coef(mod_hier_slopes)[[1]]["floor"]
ses <- (attr(ranef(mod_hier_slopes)[[1]], "postVar"))[2,2,1:n_counties]
df <- res %>% bind_cols(ses) 
colnames(df) <- c("alpha", "se")
df$county <- rownames(df)

df %>% 
  mutate(county = str_trim(county)) %>% 
  mutate(county = factor(county)) %>% 
  mutate(county = fct_reorder(county, alpha)) %>% 
  filter(county!="") %>% 
  ggplot(aes(county,alpha)) + geom_point() + coord_flip()+
  geom_errorbar(aes(ymin = alpha - se, ymax = alpha + se))
```

# Hierarchical logistic regression

## Hierarchical logistic regression

We can easily extend the idea of modeling hierarchical data to cases where our outcome of interest is a binary variable and we want to use logistic regression. 

- Recall for binary data, we have observations of our outcome $y_1, y_2, \dots, y_n$ where $y_i$ is equal to 1 if the outcome of interest occurred for observation $i$ and 0 otherwise. 
- We are interested in estimating the probability that the outcome occurs associated with one or more covariates $X_i$, i.e. $\text{Pr}(Y_i=1|X_i)$
- In usual logistic regression we model this as
$$
\text{logit Pr}(Y_i=1|X_{i1}, \dots X_{ik}) = \beta_0 + \beta_1 X_{i1} \dots + \beta_k X_{ik}
$$

## Hierarchical logistic regression

Changing the notation slightly, this is the same as 
$$
Y_i \sim \text{Bernoulli} (p_i)
$$
with 
$$
\text{logit }p_i = \beta_0 + \dots 
$$
So for hierarchical logistic regression we can model the probabilities of interest ($p_i$'s) with a hierarchical set-up

## Motivating example

Abortion outcomes in Uganda

- Data from 2018 PMA survey (via IPUMS)
- Interested in factors associated with women ever having an abortion
- Outcome of interest: 'ever had abortion (yes/no)' 
- Notes: dropping don't knows, including 'unsuccessful abortions' in 'yes'. 
- More notes: Self-reported abortion is very likely to be under-reported


# Some graphs

## Proportion by region

```{r}
library(here)
d <- read_csv(here("data/pma.csv"))

d <- d %>% filter(age>9)

props <- d %>% 
  group_by(region) %>% 
  summarize(prop_abo = mean(abortion)) %>% 
  mutate(region = fct_reorder(region, prop_abo)) 
ggplot(props, aes(region, prop_abo)) + geom_bar(stat = "identity", fill = "firebrick4") + 
  coord_flip() + 
  ylab("proportion of women surveyed who have had an abortion")+
  theme_bw(base_size = 14)
```

## Proportion by age group

```{r}
age_groups <- seq(0, 45, by = 5)
d$age_group <- as.numeric(as.character(cut(d$age, 
                   breaks= c(age_groups, Inf), 
                   labels = age_groups, 
                   right = FALSE)))

props_age <- d %>% 
  group_by(age_group) %>% 
  summarize(prop_abo = mean(abortion)) 
ggplot(props_age, aes(age_group, prop_abo)) + 
  geom_line() + 
  geom_point() + 
  labs(y = "proportion reporting ever had an abortion", x = "age group")+
  theme_bw(base_size = 14)

```

## Relationship with HDI

```{r}
prop_hdi <- d %>% 
  group_by(region, hdi) %>% 
  summarize(abo = mean(abortion)) 

ggplot(prop_hdi, aes(hdi, abo, label = region)) + 
  geom_text() +
  geom_smooth(method = "lm") + 
  labs(y = "proportion reporting ever had an abortion", x = "HDI")+
  theme_bw(base_size = 14)
  
```

## Model 1

Let's fit

$$
Y_i \sim \text{Bernoulli}(p_i)
$$
$$
\text{logit} p_i = \alpha_{j[i]} + \beta x_i
$$
$$
\alpha_j \sim N(\mu, \sigma^2)
$$
where

- $Y_i$ refers to whether or not individual $i$ has had an abortion
- The index $j$ refers to region
- $x_i$ refers to age

## Fit in R

\tiny
```{r, echo = TRUE}
mod <- glmer(abortion ~ (1|region)+age, data = d, family = "binomial")
summary(mod)
```

## Region-specific intercepts

```{r}
# pull out alphas
nregions <- length(unique(d$region))
alphas <- coef(mod)[[1]]["(Intercept)"]
alphas <- alphas %>% mutate(region = rownames(alphas)) 
ses <- attr(ranef(mod)[[1]], "postVar")[,,1:nregions]
alphas <- alphas %>% 
  mutate(se = ses) %>%
  rename(alpha = '(Intercept)') %>% 
  mutate(region = fct_reorder(region, alpha))
rownames(alphas) <- NULL

#plot

ggplot(alphas, aes(region, alpha)) + geom_point() + 
  geom_errorbar(aes(ymin = alpha - se, ymax = alpha + se)) + 
  coord_flip() + 
  theme_bw(base_size = 14) + ggtitle("Region-specific intercepts")
```

## Converting to probabilities

What is probability of abortion for a woman aged 30 in north region? The info we need:

\tiny
```{r, echo = TRUE}
alphas %>% filter(region == "north")
```

```{r, echo = TRUE}
coef(summary(mod))
```

## Model 2


$$
Y_i \sim \text{Bernoulli}(p_i)
$$
$$
\text{logit} p_i = \alpha_{j[i]} + \beta x_i
$$
$$
\alpha_j \sim N(\gamma_0 + \gamma_1 z_j, \sigma^2)
$$

where everything is as before and $z_j$ is the standardized log HDI at the region level 

## Fit in R

\tiny
```{r, echo = TRUE}
d <- d %>% mutate(log_hdi_c = (log(hdi)-mean(log(hdi)))/sd(log(hdi)))
mod2 <- glmer(abortion ~ (1|region)+age+log_hdi_c, data = d, family = "binomial")
summary(mod2)
```

## Region-specific intercepts versus HDI

```{r}
gamma0 <- coef(summary(mod2))[1,1]
gamma1 <- coef(summary(mod2))[3,1]

alphas <- coef(mod)[[1]]["(Intercept)"]
alphas <- alphas %>% mutate(region = rownames(alphas)) 
ses <- attr(ranef(mod)[[1]], "postVar")[,,1:nregions]
alphas <- alphas %>% 
  mutate(se = ses) %>%
  rename(alpha = '(Intercept)') %>% 
  mutate(region = fct_reorder(region, alpha))
rownames(alphas) <- NULL

hdi <- d %>% group_by(region, hdi) %>% 
  slice(1) %>% 
  select(region, hdi)

yhat <- hdi %>% 
  left_join(alphas) %>% 
  ungroup() %>% 
  mutate(log_hdi_c = (log(hdi)-mean(log(hdi)))/sd(log(hdi))) %>% 
  mutate(y_hat = alpha + gamma1*log_hdi_c)


ggplot(yhat, aes(log_hdi_c, y_hat)) + 
  geom_point()+
  geom_errorbar(aes(ymin = y_hat - 2*se, ymax = y_hat + 2*se)) +
   geom_abline(intercept = gamma0, slope = gamma1) +
  ylab("logit proportion of abortion") + xlab("standardized log HDI") + 
  theme_bw(base_size = 14)
```

## Age as factor

\tiny
```{r, echo = TRUE}
d <- d %>% mutate(age_group = factor(age_group)) %>% 
  mutate(age_group = fct_relevel(age_group, "20", after = 0))
mod3 <- glmer(abortion ~ (1|region)+age_group+log_hdi_c, data = d, family = "binomial")
coef(summary(mod3))
```
# Non-nested hierarchies

## Non-nested hierarchies

We can extend hierarchical structure to be on more than one variable, e.g. by region and age group.

$$
Y_i \sim \text{Bernoulli}(p_i)
$$
$$
\text{logit} p_i = \beta_0 + \alpha_{j[i]} + \eta_{k[i]}
$$
$$
\alpha_j \sim N(0, \sigma_j^2)
$$
and 
$$
\eta_k \sim N(0, \sigma_k^2)
$$
where $k$ refers to age group of individual $i$

- Notice now everthing is centered at zero and we have a global intercept

## Rewrite model



## Adding non-nested hierarchies in R

\tiny
```{r, echo = TRUE}
mod4 <- glmer(abortion ~ (1|region)+(1|age_group), data = d, family = "binomial")
summary(mod4)
```

## Summary

- Can extend hierarchical models to be varying slopes as well as intercepts
- Can use hierarchical models in a logistic regression context where the outcome of interest is binary
- In general, useful when you have small sample sizes in some cells (i.e. some population subgroups)
- Standardizing covariates (minusing meand and dividing by standard deviation) is useful for interpretation (sometimes) but also to fit models

# Lab - R Markdown 

