---
title: "Formatting examples"
author: "Monica Alexander"
date: "26/03/2021"
output: 
  pdf_document:
    number_sections: true
fontsize: 12pt
header-includes:
  - \usepackage{setspace} \onehalfspacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

# Introduction

Here is my introduction, I want to talk about formatting. I'm writing some more text so I have text on multiple lines and so you can see that it's single spaced.

Here is an example of *italics*.

Here is an example of **bolding**.

Trying to \text{\underline{underline}}.


## Overview

- Hiding code, message, warnings
- Numbering sections
- Fontsize, spacing
- Underlining, bolding, italicizing
- Writing math
- Table formatting

# Data

Here is my data:


```{r}
library(tidyverse)
library(here)
d <- read_csv(here("data/pma.csv"))
head(d)
```


## Summary statistics


```{r}
prop_by_region <- d %>% 
  group_by(region) %>% 
  summarize(prop_abortion = mean(abortion)) %>% 
  arrange(prop_abortion)
prop_by_region
```

KableExtra package

```{r}
library(kableExtra)

prop_by_region %>% 
  kable(booktabs = TRUE, caption = "My caption", col.names = c("Region", "Proportion of women had abortion")) %>%
  kable_styling(latex_options = "striped" )

```


## Bivariate relationships

# Methods

## Writing math

The way that we write math in Rmarkdown essentially you write within dollar signs. 

Note: one dollar sign is same line. Define $Y_i$ to be whether or not woman $i$ has ever had an abortion. 

Two dollars signs is new line and indented. This is for when you want to write an equation. 

$$
Y_i = \beta_0 + \beta_1 X_i
$$
Keep typing below

Things to note:

- Subscripts, write after underscore $Y_i$
- If you want more than one subscript, the subscript has to go in curly brackets. $X_{i1}$
- Greek letters: start with backslash and then write the name of the letter. $\beta$, $\alpha$, $\sigma$
- Subscripts on greek: $\beta_0$
- Superscript: $\sigma^2$

Example of a hierarchical model:

$$
Y_i \sim N(\alpha_{j[i]}, \sigma^2)
$$
with

$$
\alpha_j \sim N(\mu_{\alpha}, \sigma^2_{\alpha})
$$

Logistic regression:

$$
\text{logit }Y_i = \beta_0 + \beta_1X_i
$$


Numbered lists:

1. hliu
2. iugouygdlyugs

# Discussion

# Reference

