---
title: 'MRP example: self-reported abortion in Uganda'
author: "Monica Alexander"
date: "21/04/2021"
output: 
  pdf_document:
    number_sections: true
    toc: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This document goes through a worked example of multilevel regression and post-stratification to get estimates of self-reported abortion incidence in Uganda. The data used are the 2018 PMA survey and the 2014 Census. Both datasets were obtained through IPUMS. 

# Load in data and tidy up 

Load in the packages:

```{r}
library(tidyverse)
library(here)
library(brms)
library(tidybayes)
```

## PMA

Load in the PMA data and make an age category variable:

```{r}
d <- read_csv(here("data/pma.csv"))

d <- d %>% filter(age>14, age<50)

age_groups <- seq(15, 45, by = 5)
d$age_group <- as.numeric(as.character(cut(d$age, 
                                           breaks= c(age_groups, Inf), 
                                           labels = age_groups, 
                                           right = FALSE)))

d <- d %>% mutate(age_group = factor(age_group)) %>% 
  mutate(age_group = fct_relevel(age_group, "35", after = 0))
```

Create a new `region_2` variable with slightly bigger regions (that match the census data):

```{r}
d <- d %>% mutate(region_2 = case_when(region == "north"|region== "karamoja"|region == "west nile" ~ "northern",
                                       region == "eastern" ~ "eastern",
                                       region == "central 1"|region == "central 2"|region == "east central"|region == "kampala" ~ "central",
                                       region == "western"|region == "south west" ~ "western"))

```

## Census

Load in the data, select the columns we want and make an age category variable:

```{r}
dc <- haven::read_dta(here("data/uganda_census.dta"))

dc <- dc %>% 
  select(regnug, perwt, age, marst, edattain) %>% 
  mutate(regnug = as_factor(regnug),
         marst = as_factor(marst),
         edattain = as_factor(edattain))

dc <- dc %>% filter(age>14, age<50)


dc$age_group <- as.numeric(as.character(cut(dc$age, 
                                           breaks= c(age_groups, Inf), 
                                           labels = age_groups, 
                                           right = FALSE)))

dc <- dc %>% mutate(age_group = factor(age_group)) %>% 
  mutate(age_group = fct_relevel(age_group, "35", after = 0))

```

## Recode education and marital status

The PMA and census have different education and marital status categories. Let's recode so they are the same:

```{r}

# EDUCATION

#table(d$educattgen)
#table(dc$edattain)


d <- d %>% mutate(educ = case_when(educattgen=="never attended"~"less than primary",
                                   educattgen=="primary/middle school" ~"primary",
                                   educattgen=="secondary/post-primary"|educattgen=="tertiary/post-secondary" ~ "more than primary",
                                   TRUE ~ "NA")) %>% 
  filter(educ != "NA", marstat!="no response or missing")

dc <- dc %>% mutate(educ = case_when(edattain=="less than primary completed"~"less than primary",
                                     edattain=="primary completed"~ "primary",
                                     edattain=="secondary completed"|edattain=="university completed" ~ "more than primary",
                                     TRUE ~ "NA")) %>% 
  filter(educ !="NA")

## MARITAL STATUS

#table(d$marstat)
#table(dc$marst)

d <- d %>% mutate(marital = case_when(marstat == "never married" ~ "single/never married",
                                      marstat == "currently living with partner"| marstat == "currently married" ~ "married/in union",
                                      marstat== "divorced or separated" ~ "divorced/separated",
                                      marstat== "widow or widower"~ "widowed",
                                      TRUE ~ "NA"))

dc <- dc %>% mutate(marital = case_when(marst == "single/never married" ~ "single/never married",
                                        marst == "married/in union" ~ "married/in union",
                                        marst== "separated/divorced/spouse absent" ~ "divorced/separated",
                                        marst== "widowed"~ "widowed",
                                      TRUE ~ "NA")) %>% 
  filter(marital != "NA")
                                
```


# Plot data

Lets calculate the census population by key subgroups:

```{r}
census_counts <- dc %>% 
  group_by(regnug, marital, educ, age_group) %>% 
  summarize(n = sum(perwt)) %>% 
  filter(age_group!="45") %>% 
  rename(region_2 = regnug)
```



We can get an idea of differences in population distributions by plotting the PMA and census proportions by different variables:

```{r}
d %>% 
  group_by(age_group) %>% 
  tally() %>% 
  mutate(pma = n/sum(n)) %>% 
  left_join(census_counts %>% 
              group_by(age_group) %>% 
              summarize(n = sum(n)) %>% 
              mutate(census = n/sum(n)) %>% 
              select(-n)) %>% 
  mutate(age_group = as.character(age_group)) %>% 
  arrange(age_group) %>% 
  pivot_longer(pma:census) %>% 
  ggplot(aes(age_group, value, fill = name)) + geom_bar(stat = "identity", position = 'dodge')+
  theme_bw(base_size = 14) + 
  labs( x = "age", y = "proportion", title = "Proportion by age group")+
  scale_fill_brewer(palette = "Set1", name = "source")


d %>% 
  group_by(educ) %>% 
  tally() %>% 
  mutate(pma = n/sum(n)) %>% 
  left_join(census_counts %>% 
              group_by(educ) %>% 
              summarize(n = sum(n)) %>% 
              mutate(census = n/sum(n)) %>% 
              select(-n)) %>% 
  mutate(educ = fct_relevel(educ, "more than primary", after = 2)) %>% 
  pivot_longer(pma:census) %>% 
  ggplot(aes(educ, value, fill = name)) + geom_bar(stat = "identity", position = 'dodge')+
  theme_bw(base_size = 14) + 
  labs( x = "education", y = "proportion", title = "Proportion by education")+
  scale_fill_brewer(palette = "Set1", name = "source")


d %>% 
  group_by(region_2) %>% 
  tally() %>% 
  mutate(pma = n/sum(n)) %>% 
  left_join(census_counts %>% 
              group_by(region_2) %>% 
              summarize(n = sum(n)) %>% 
              mutate(census = n/sum(n)) %>% 
              select(-n)) %>% 
  pivot_longer(pma:census) %>% 
  ggplot(aes(region_2, value, fill = name)) + geom_bar(stat = "identity", position = 'dodge')+
  theme_bw(base_size = 14) + 
  labs( x = "region", y = "proportion", title = "Proportion by region")+
  scale_fill_brewer(palette = "Set1", name = "source")



d %>% 
  group_by(marital) %>% 
  tally() %>% 
  mutate(pma = n/sum(n)) %>% 
  left_join(census_counts %>% 
              group_by(marital) %>% 
              summarize(n = sum(n)) %>% 
              mutate(census = n/sum(n)) %>% 
              select(-n)) %>% 
  pivot_longer(pma:census) %>% 
  ggplot(aes(marital, value, fill = name)) + geom_bar(stat = "identity", position = 'dodge')+
  theme_bw(base_size = 14) + 
  labs( x = "marital status", y = "proportion", title = "Proportion by marital status")+
  scale_fill_brewer(palette = "Set1", name = "source")+coord_flip()
```


# Multilevel regression

Let's run a logistic regression of whether or not an individual reported ever having an abortion with covariates education, age group, and region modelled hierarchically

## Run the model

```{r}
mod <- brm(abortion ~ (1|region_2)+educ+ marital +age_group, data = d, family = "bernoulli")
summary(mod)

# optional: cf to lme4
#summary(lme4::glmer(abortion ~ (1|region_2)+marital+educ+ age_group, data = d, family = "binomial"))
```

## Get estimated proportions of abortion incidence

Now we need to get the estimated proportions of abortion incidence by each subpopulation of interest. (i.e. by region, marital status, education and age group). First, initiate a tibble with all possible combinations:

```{r}
regions <- unique(d$region_2)
marital_groups <- unique(d$marital)
educ_groups  <- unique(d$educ)

pred_df <- tibble(region_2 = NA, marital = NA, educ = NA, age_group = NA) %>% 
  tidyr::complete(region_2 = regions, 
                  marital = marital_groups,
                  educ = educ_groups, 
                  age_group = age_groups) %>% 
  mutate_all(.funs = funs(as_factor(.))) %>% 
  drop_na() %>% 
  mutate(age_group = fct_relevel(age_group, "35", after = 0)) %>% 
  filter(age_group != "45")
```

We can use the `fitted_draws` function to obtain posterior samples of the estimated proportion of women reporting an abortion by each group:

```{r}
pred_probs_draws <- mod %>% 
  fitted_draws(pred_df  %>% 
                    select(region_2:age_group) %>% 
                    mutate(age_group = factor(age_group, levels = age_groups[-length(age_groups)])))
```


# Post-stratify the predicted proportions

For the post-stratfication, we need a data frame that tells us how many women in the census are in each subgroup of interest. Let's join the census counts to the `pred_probs_draws` table above, replacing any NAs with 0's:

```{r}
pred_probs_draws <- pred_probs_draws %>% 
  left_join(census_counts) %>% 
  replace_na(replace = list(n = 0))
```

For each group and draw, we can calculate the estimated number of women reporting an abortion:

```{r}
pred_probs_draws <- pred_probs_draws %>% 
  mutate(n_abo = n*.value)
```

We can now use this as a basis of getting estimates by any group of interest. For example, the national estimate and lower/upper bounds is

```{r}
pred_probs_draws %>% 
  group_by(.draw) %>% 
  summarize(prop_abo = sum(n_abo)/sum(n)) %>% 
  ungroup() %>% 
  summarize(prop_abo_group = median(prop_abo), 
            lower = quantile(prop_abo, 0.1),
            upper = quantile(prop_abo, 0.9))
```

By age group:

```{r}
prop_by_age <- pred_probs_draws %>% 
  group_by(.draw, age_group) %>% 
  summarize(prop_abo = sum(n_abo)/sum(n)) %>% 
  group_by(age_group) %>% 
  summarize(prop_abo_group = median(prop_abo), 
            lower = quantile(prop_abo, 0.1),
            upper = quantile(prop_abo, 0.9))

prop_by_age
```

By marital status:

```{r}
prop_by_marital <- pred_probs_draws %>% 
  group_by(.draw, marital) %>% 
  summarize(prop_abo = sum(n_abo)/sum(n)) %>% 
  group_by(marital) %>% 
  summarize(prop_abo_group = median(prop_abo), 
            lower = quantile(prop_abo, 0.1),
            upper = quantile(prop_abo, 0.9))

prop_by_marital
```

By education:

```{r}
prop_by_educ <- pred_probs_draws %>% 
  group_by(.draw, educ) %>% 
  summarize(prop_abo = sum(n_abo)/sum(n)) %>% 
  group_by(educ) %>% 
  summarize(prop_abo_group = median(prop_abo), 
            lower = quantile(prop_abo, 0.1),
            upper = quantile(prop_abo, 0.9))

prop_by_educ
```

By region:

```{r}
prop_by_region <- pred_probs_draws %>% 
  group_by(.draw, region_2) %>% 
  summarize(prop_abo = sum(n_abo)/sum(n)) %>% 
  group_by(region_2) %>% 
  summarize(prop_abo_group = median(prop_abo), 
            lower = quantile(prop_abo, 0.1),
            upper = quantile(prop_abo, 0.9))

prop_by_region
```


# Plotting and comparing estimates

We can compare the MRP estimates with raw estimates from the PMA survey and also estimates using normal post-stratification

## Calculating raw estimates and post-stratified estimates

Calculate the raw estimates from the survey:

```{r}
pma_cells <- d %>% 
  group_by(region_2, educ, marital, age_group) %>% 
  summarize(
            n_sample = n(), n_abo_sample = sum(abortion),
            prop_abo = n_abo_sample/n_sample)
```

We can then combine these with census data to get post-stratified counts:

```{r}
pma_cells <- pma_cells %>% 
  left_join(census_counts) %>% 
  mutate(n_abo = prop_abo*n) %>% 
  replace_na(replace = list(n = 0,n_abo = 0)) %>% 
  ungroup()
```


## Join all estimates together

Calculate raw and postratified estimates by different subpopulations and join them to the MRP estimates. By age group:

```{r}
prop_by_age <- prop_by_age %>% 
  mutate(type = "mrp") %>% 
  rename(point = prop_abo_group) %>% 
  bind_rows(pma_cells %>% group_by(age_group) %>%summarize(poststrat = sum(n_abo)/sum(n)) %>% 
  left_join(pma_cells %>% group_by(age_group) %>%summarize(raw = sum(n_abo_sample)/sum(n_sample))) %>% 
  pivot_longer(-age_group, names_to = "type", values_to = "point")) %>% 
  arrange(age_group)
```

By education:

```{r}
prop_by_educ <- prop_by_educ %>% 
  mutate(type = "mrp") %>% 
  rename(point = prop_abo_group) %>% 
  bind_rows(pma_cells %>% group_by(educ) %>%summarize(poststrat = sum(n_abo)/sum(n)) %>% 
  left_join(pma_cells %>% group_by(educ) %>%summarize(raw = sum(n_abo_sample)/sum(n_sample))) %>% 
  pivot_longer(-educ, names_to = "type", values_to = "point")) %>% 
  arrange(educ)
```

By marital status:

```{r}
prop_by_marital <- prop_by_marital %>% 
  mutate(type = "mrp") %>% 
  rename(point = prop_abo_group) %>% 
  bind_rows(pma_cells %>% group_by(marital) %>%summarize(poststrat = sum(n_abo)/sum(n)) %>% 
  left_join(pma_cells %>% group_by(marital) %>%summarize(raw = sum(n_abo_sample)/sum(n_sample))) %>% 
  pivot_longer(-marital, names_to = "type", values_to = "point")) %>% 
  arrange(marital)
```

By region:

```{r}
prop_by_region <- prop_by_region %>% 
  mutate(type = "mrp") %>% 
  rename(point = prop_abo_group) %>% 
  bind_rows(pma_cells %>% group_by(region_2) %>%summarize(poststrat = sum(n_abo)/sum(n)) %>% 
  left_join(pma_cells %>% group_by(region_2) %>%summarize(raw = sum(n_abo_sample)/sum(n_sample))) %>% 
  pivot_longer(-region_2, names_to = "type", values_to = "point")) %>% 
  arrange(region_2)
```

## Plot!

By age group:

```{r}
prop_by_age %>% 
  ggplot(aes(age_group, point, color = type)) + geom_point(size = 2) + 
  geom_errorbar(aes(ymin = lower, ymax = upper), width = NA, alpha = 0.5) + 
  theme_bw(base_size = 14) +
  labs(title = "Abortion prevalence by age group", x = "age", y = "proportion of women reporting abortions")
```

By education: 

```{r}
prop_by_educ %>% 
  mutate(educ = fct_relevel(educ, "more than primary", after = 2)) %>% 
  ggplot(aes(educ, point, color = type)) + geom_point(size = 2) + 
  geom_errorbar(aes(ymin = lower, ymax = upper), width = NA, alpha = 0.5) + 
  theme_bw(base_size = 14) +
  labs(title = "Abortion prevalence by education", x = "education level", y = "proportion of women reporting abortions")
```

By marital status

```{r}
prop_by_marital %>% 
  mutate(marital = factor(marital, c("single/never married", "married/in union", "divorced/separated", "widowed"))) %>% 
  ggplot(aes(marital, point, color = type)) + geom_point(size = 2) + 
  geom_errorbar(aes(ymin = lower, ymax = upper), width = NA, alpha = 0.5) + 
  theme_bw(base_size = 14) +
  labs(title = "Abortion prevalence by marital status", x = "", y = "proportion of women reporting abortions")+
  coord_flip()
```

By region

```{r}
prop_by_region %>% 
  ggplot(aes(region_2, point, color = type)) + geom_point(size = 2) + 
  geom_errorbar(aes(ymin = lower, ymax = upper), width = NA, alpha = 0.5) + 
  theme_bw(base_size = 14) +
  labs(title = "Abortion prevalence by region", x = "", y = "proportion of women reporting abortions")
```

