---
title: 'Week 3: Linear regression'
author: "Monica Alexander"
date: "26/01/2021"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# By the end of this lab you should know 

- How to use `lm` to get estimated coefficients and $R^2$
- How to calculate an estimated outcome (Y) based on a particular value of an independent variable (X) and estimated regression coefficients
- How to plot a fitted SLR line on a scatter plot
- How to extract fitted values and residuals from `lm` object
- How to extract coefficient and standard error estimates from `lm` summary object
- How to find critical t values
- How to calculate confidence intervals


# Read in, prepare, plot the data

We will be using the country indicators dataset again, and exploring the relationship between the total fertility rate (TFR) and child mortality in 2017. 

NOTE: If you are having trouble with the here package, I suggest not using it and just putting in the entire file path.

```{r}
library(tidyverse)
library(here)
country_ind <- read_csv(here("data/country_indicators.csv"))

# NOTE if you are having trouble with the 'here' package
# don't use it and just type in the whole file path. 
# E.g. Monica's would be
# country_ind <- read_csv("~/src/soc252/data/country_indicators.csv")
```


## Filter to just be 2017

```{r}
country_ind_2017 <- country_ind %>% filter(year==2017)
```

## Look at the observed relationship between TFR and child mortality

```{r}
ggplot(country_ind_2017, aes(tfr, child_mort)) + 
  geom_point() + 
  ylab("Child mortality") + 
  xlab("TFR") + 
  ggtitle("Child mortality versus TFR, 2017")
```

## Question

Alter the code above to make a plot title and make the X and Y axes more readable. 

# Estimating SLR using `lm`

We don't have to calculate the regression coefficients or $R^2$ 'by hand', we can just use the `lm` function. ('lm' stands for 'linear models').

The main arguments are

- The formula, which is written in the form `y~x` 
- The data frame that contains the variables 

Fit our SLR:

- Y = child mortality
- X = TFR

```{r}
childmort_tfr_model <- lm(formula = child_mort~tfr, # formula, so this tells lm what our model is
                          data = country_ind_2017) # where the data come from
```

Print out the summary:

```{r}
summary(childmort_tfr_model)
```
BAD PRACTICE: 

- losing accuracy
- prone to human error
- not reproducible, and doesn't automatically update

```{r}
b1_hat <- 20.3581
```

```{r}
gss <- read_csv(here("data/gss.csv"))

# FIND the proportion of people who have non-missing age_first_child

# total number of respondents

n <- nrow(gss)
n

# the number of non-missing values

n_non_missing <- gss %>% select(age_first_child) %>% 
  summarise(sum(!is.na(age_first_child))) %>% 
  pull()

n_non_missing


# proportion

n_non_missing/n #fine, this is good

prop <- n_non_missing/n


# OR

13767/20602 # bad practice


```




## Extract results

To extract coefficients from the model output, use the `coef()` function

```{r}
coef(childmort_tfr_model)
```

Can assign these to variables by indexing the relevant number:

```{r}
beta_0 <- coef(childmort_tfr_model)[1] # the [1] means get the first item
beta_1 <- coef(childmort_tfr_model)[2] 

beta_0
```

To extract the value of $R^2$, use

```{r}
summary(childmort_tfr_model)[["r.squared"]]
```

Use `names()` function to see what values are within an object

```{r}
names(summary(childmort_tfr_model))
```



## Question

What is the estimated child mortality for a country with a TFR of 5?

$Y = \beta_0 + \beta_1X$

- $X$ = TFR
- $Y$ = child mortality

Based on our estimation, what is $\hat{Y}$ when $X = 5$?

Need: estimates of beta_0 and beta_1


```{r}
beta_0
beta_1
x <- 5

beta_0+beta_1*x

```




# Plotting the fitted line on a scatter plot

To visualize our fitted line we can add to our plot from before using `geom_abline`

```{r}
ggplot(country_ind_2017, aes(tfr, child_mort)) + 
  geom_point() + 
  geom_abline(intercept = beta_0, slope = beta_1, color = "red") # adds a line with a specified intercept and slope
```


# Extract fitted values and residuals

To extract the fitted values of child mortality:

```{r}
Yhat <- fitted(childmort_tfr_model)
Yhat

```

To extract the residuals:

ehat = Y - Yhat

```{r}
ehat <- resid(childmort_tfr_model)
ehat
```

## Questions

- Calculate Ybar (i.e. the mean of `life expectancy`)
- Hence calculate SSM
- Calculate SSR
- Hence calculate SST and $R^2$


```{r}
summary(childmort_tfr_model)[["r.squared"]]
```

```{r}
Ybar <- country_ind_2017 %>% 
  summarise(mean(child_mort)) %>% 
  pull() # pulls out the value to just be one number 

Ybar
```
SSM : sum of the fitted values - mean all squared

```{r}
SSM <- sum((Yhat - Ybar)^2)

SSM

```

SSR: sum of observed - fitted all squared

```{r}
# pull out our "Y" values (which are observed child mortality)

Y <- country_ind_2017 %>% 
  select(child_mort) %>% 
  pull() # pulls out our set of values

Y

SSR <- sum((Y - Yhat)^2)
SSR
```


SST = SSM + SSR
R^2 = SSM/SST

```{r}
SST <- SSM + SSR
SST

SSM/SST

```


# Extracting standard errors of coefficients

Easiest to do based on extracting info out of the `summary` object. To look at everything that's contained in the summary object use `names`

```{r}
summary_mod <- summary(childmort_tfr_model)

names(summary_mod)
```

We want to look at coefficients. We use the dollar sign `$` notation here to extract a particular piece of the model object. 

```{r}
summary_mod[["coefficients"]]

#OR

summary_mod$coefficients
```

Notice this is a matrix with 2 rows and 4 columns. To get $\hat{\beta}_1$, it's the 2nd row and 1st column:

```{r}
b1_hat <- summary_mod$coefficients[2,1]
b1_hat
```

To get the standard error, it's the 2nd row and 2nd column:

```{r}
se_b1_hat <- summary_mod$coefficients[2,2]
se_b1_hat
```

# Calculate confidence interval

We will calculate a 95% confidence interval i.e. $\alpha = 0.05$.

First, to get the critical value, we use the `qt` function

```{r}
alpha <- 0.05 # define our alpha (type 1 error rate)
n <- nrow(country_ind_2017) # number of observations
df <- n-2 # degrees of freedom (n - (k+1))

# take absolute value to ensure it's positive
# p = alpha/2 because two sided 
t_alpha <- abs(qt(p = alpha/2, df = df))
t_alpha
```

The formula for the confidence interval is 
$$\hat{\beta}_{1} \pm\left(t_{\alpha} \times s e\left(\hat{\beta}_{1}\right)\right)$$

Where the symbol $\pm$ means you do both plus and minus to get two different values, which forms the lower and upper bound of your confidence interval

## Questions
- Calculate the 95% confidence interval for $\hat{\beta}_{1}$
- Interpret this confidence interval


```{r}
# lower bound 

b1_hat - t_alpha*se_b1_hat

# upper bound 

b1_hat + t_alpha*se_b1_hat

b1_hat

```

Estimate for beta1_hat = 20.36 (95% CI: [18.66, 22.05]). Interpretation: if we could get repeated samples of child mortality and TFR over and over, and calculated the confidence interval for each sample, the 95% of those intervals would contain the true value of beta1_hat. 



# Multiple linear regression

Running MLR in R is an easy extension of SLR. Here are some practice questions using the `lego_towers` dataset. This dataset shows observations of the time (in seconds) it took my toddler to build a lego tower, the number of blocks given to him, and the number of other distractions present. 

Reading in the data:

```{r}
lego <- read_csv(here("data/lego_towers.csv"))
lego
```
## Questions

1. Make a scatter plot of time versus blocks

```{r}
ggplot(data = lego, aes(x = blocks, y = time)) + 
  geom_point()
```


2. Make a scatter plot of time versus distractions

```{r}
ggplot(data = lego, aes(x = distractions, y = time)) + 
  geom_point()
```



3. Based on 1 and 2, what do you expect the magnitude and sign (direction) of $\hat{\beta_1}$ and $\hat{\beta_2}$ to be?

X1 = blocks
X2 = distractions
Y  = time

- beta1_hat, beta2_hat both positive
- expect beta1_hat (blocks) to have higher magnitude

4. Fit the above model using `lm`

```{r}
lego_mod <- lm(time~blocks+distractions, data = lego)
summary(lego_mod)
```


5. Interpret $\hat{\beta_1}$ and $\hat{\beta_2}$

- beta1: for every additional block, the time taken to build the tower increases by 4.7 seconds, holding distractions constant. 
- beta2: for every additional distraction, the time taken to buld the tower increases by 1.1 seconds, holding the number of blocks constant. 


6. Using `mutate` create a new variable called `blocks_3` which is the number of blocks minus 3

```{r}
lego <- lego %>% 
  mutate(blocks_3 = blocks - 3)

lego

```


7. Refit the model using `lm` where $X_{i1}$ is now `blocks_3`

```{r}
lego_mod_2 <- lm(time~blocks_3+distractions, data = lego)
summary(lego_mod_2)
```

Notice

- beta1hat and beta2hat have not changed
- Intercept value has changed. 



8. Interpret $\hat{\beta_0}$

Beta0hat: time taken to build a tower when `blocks_3` and distractions is 0. 
When `blocks_3` = 0, blocks = 3. 

More helpful interpretation: with 3 blocks and no distractions, the average time to build a tower is 5.8 seconds. 



REMOVE STUFF:

- remove one variable: use `rm()`


```{r}
rm(beta_0)

beta_0
```

- Want to get rid of everything in your environment:
    + save your work and restart RStudio
    + broom button top RH corner



life expectancy and tfr over time

```{r}
country_ind_subset <- country_ind %>% 
  filter(country=="Australia"|country == "Kenya")

ggplot(data = country_ind_subset, aes(x = tfr, y = life_expectancy, color = year)) + 
  geom_point() + 
  facet_wrap(~country)

country_ind_subset <- country_ind_subset %>% 
  mutate(ratio = life_expectancy/tfr)

ggplot(country_ind_subset, aes(year, ratio, color = country)) + 
  geom_line() + 
  geom_point()
```








