---
title: 'Week 9: Logistic regression'
author: "Monica Alexander"
date: "09/02/2021"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

We are going to look at the association of income (low income or not), age and education in the GSS. 

# Data

Load in the GSS


```{r}
library(tidyverse)
library(here)
gss <- read_csv(here("data/gss.csv"))
```

# Create binary outcome variable

The income variable in the GSS is categorical. You can look at different categories by using the `unique` function:

```{r}
unique(gss$income_respondent)
```

For the purposes of our logistic regression, let's define a binary variable called "low income" that is equal to 1 if a respondent is in the less than $25,000 category and 0 otherwise. 

```{r}
# if the income is in this "Less than $25,000" category, set low_income to be 1, else set low_income to be 0.

gss <- gss %>% 
  mutate(low_income = ifelse(income_respondent=="Less than $25,000", 1, 0))

gss %>% 
  select(income_respondent, low_income)

```

NOTE: this is not the only possible way of studying income as a binary outcome. For example, you may be more interested in whether or not a respondent has high income, and define a binary variable called `high_income` if the respondent has more than \$125,000. Additionally, you can define a binary outcome based on more than one category. For example, if we were interested in whether respondents earned more or less than \$50,000, we could define the following:

```{r}

# "|" = or

gss <- gss %>% 
  mutate(income_less_than_50k = ifelse(income_respondent=="Less than $25,000"|
                                         income_respondent=="$25,000 to $49,999", 1, 0))
```

Note that the vertical bar "|" means "or" so you read the above code as "if the income of the respondent is less than \$25,000 OR between \$25,000 and \$49,000 then the new variable `income_less_than_50k` is equal to 1, otherwise it is equal to 0."

# Logistic regression

Now run a logistic regression with dependent variable `low_income` and independent variable age. 

```{r}
# two differences: use glm not lm
# need to specify a family, for logistic regression, this will always be binomial

mod <- glm(low_income ~ age, family = "binomial", data = gss)
summary(mod)
```

Remember that the coefficients are on the 'log-odds' scale. To convert to the odds scale, you can exponentiate:

```{r}
coef(mod)
exp(coef(mod))
```


## Question

Interpret $\beta_1$ and $\exp \beta_1$

- beta_1 = -0.006: Increasing age by one year decreases the log odds of low income by 0.006.
- exp(beta_1) = 0.99. Increasing the age by one year decreases the odds of low income by (1-0.99) = 1%.


# Regression with age groups

using age as a quantitative variable as above assumes that the association with income and age is always constant: that is, the probability of low income always decreases with increased age. But do we believe this? It might be the case that the likelihood of low income changes over age profiles. To investigate this, we can define a age group categorical variable and run a regression with this variable. 

First define a 10 year age group variable:

```{r}
age_groups <- seq(10, 80, by = 10)

gss$age_group <- as.character(cut(gss$age, 
                   breaks= c(age_groups, Inf), 
                   labels = age_groups, 
                   right = FALSE))


gss %>% 
  select(age, age_group)
```

Now run a regression:

```{r}
mod_2 <- glm(low_income ~ age_group, family = "binomial", data = gss)
summary(mod_2)
```

## Questions

- What is the reference category? Age group == 10-20
- Interpret the coefficients on age group = 20 and age group = 50. What does this suggest?

```{r}
exp(coef(mod_2))
1- exp(coef(mod_2))
```

age group 20: the odds of low income for 20-30 years is (1-0.015) = 98% less than the odds of low income for 10-20 years olds. 

age group 50: the odds of low income for 50-60 year olds is (1-0.006) = 99% less than the odds of low income for 10-20 years olds. 


# Changing reference category

Now rerun the regression based on the re-leveled age group:

```{r}

# factor relevel
# redefining our age group variable, saying that the category "30" should be first

gss <- gss %>% 
  mutate(age_group = fct_relevel(age_group, "30", after = 0))

mod_3 <- glm(low_income ~ age_group, family = "binomial", data = gss)
summary(mod_3)
```
## Questions

- Interpret the coefficients on age group = 20 and age group = 50. Why do these differ?

```{r}
exp(coef(mod_3))
```

- Age group 20 and 50 both have odds ratios of > 1
- Thta means that compared to reference category (30-40) these groups have a higher odds of low income
    + age group 20: the odds of low income are 3.14 times the oddss of low income for 30-40 years
    + age group 50: the odds of low income ae 28% higher than the odds of low income for 30-40 year olds (1.28 x)

# Further exercises

- Rerun the regression above (mod_3) with `educ_cat` as an additional explanatory variable. What is the reference category? Interpret some of the results. 

```{r}
unique(gss$educ_cat)
```

```{r}
mod_4 <- glm(low_income~age_group+educ_cat, data = gss, family = "binomial")
summary(mod_4)
```

Reference is 3-40 year olds with a bachelor's degree.

- Compared to 30-40 year olds with a bachelor's degree, 30-40 year olds with a high school certificate only are more likely to have low income (the odds of low income is 2.2x higher)

```{r}
exp(coef(mod_4))
```


- Change the reference category for education to be high school and rerun the above regression. 
```{r}

gss <- gss %>% 
  mutate(educ_cat = fct_relevel(educ_cat, "high school", after=0))

mod_5 <- glm(low_income~age_group+educ_cat, data = gss, family = "binomial")
summary(mod_5)
```

- Reference category: 30-40 year olds with high school certificate only




