---
title: 'Week 7: Interactions and Multinomial regression'
author: "Monica Alexander"
date: "04/03/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

We are going to go through two of the models mentioned in the lecture. 

Packages:

```{r}
library(tidyverse)
library(here)
library(nnet) # for multinomial
```


# Interactions

Using the country indicators dataset again.

```{r}
country_ind <- read_csv(here("data/country_indicators.csv"))
```

Question of interest: in 2017, how is TFR associated with life expectancy and whether or not a country is in a developed region?

Filter the data and create and indicator variable:

```{r}
country_ind_2017 <- country_ind %>% 
  filter(year==2017) %>% # filter to just have 2017 
  mutate(dev_region = ifelse(region=="Developed regions", "yes", "no")) # create a binary variable that indicates whether or not a country is in developed region

country_ind_2017
```

Run a model with interaction:

```{r}
mod <- lm(tfr ~ life_expectancy + dev_region + life_expectancy*dev_region, data = country_ind_2017)
summary(mod)
```

## Visualizing interactions

DON'T DO IT THIS WAY:

Grab coefficients:

```{r}
intercept_non_dev <- coef(mod)[[1]]
slope_non_dev <- coef(mod)[[2]]
intercept_dev <- intercept_non_dev+ coef(mod)[[3]]
slope_dev <- slope_non_dev+ coef(mod)[[4]]
```

Also need min and max life expectancies by region

```{r}
min_max <- country_ind_2017 %>% 
  group_by(dev_region) %>% 
  summarize(min = min(life_expectancy), max = max(life_expectancy))
min_max
```


Plot:
```{r}
ggplot(aes(life_expectancy, tfr, color = dev_region), data = country_ind_2017) + 
  geom_point() + 
  ggtitle("TFR versus life expectancy, by region") + 
  ylab("TFR") + xlab("Life expectancy") + 
  scale_color_brewer(name = "Developed region", palette = "Set1")+
  geom_segment(aes(x = min_max$min[1], xend = min_max$max[1],
                   y = intercept_non_dev+ slope_non_dev*min_max$min[1], 
                   yend = intercept_non_dev+ slope_non_dev*min_max$max[1])) +
  geom_segment(aes(x = min_max$min[2], xend = min_max$max[2],
                   y = intercept_dev+ slope_dev*min_max$min[2], 
                   yend = intercept_dev+ slope_dev*min_max$max[2]), color = "blue")
```


DO IT THIS WAY: geom_smooth()

A quicker way:

```{r}
ggplot(aes(life_expectancy, tfr, color = dev_region), data = country_ind_2017) + 
  geom_point() + 
  geom_smooth(method = "lm") + # fit line of best using linear regression
  ggtitle("TFR versus life expectancy, by region") + 
  ylab("TFR") + xlab("Life expectancy") + 
  scale_color_brewer(name = "Developed region", palette = "Set1")
```
Let's compare this to a model with no interaction effects.

```{r}
mod2 <- lm(tfr~life_expectancy+dev_region, data = country_ind_2017)
summary(mod2)
```

Pulling out the coefficients 

```{r}
intercept_non_dev <- coef(mod2)[[1]]
slope <- coef(mod2)[[2]]
intercept_dev <- intercept_non_dev+ coef(mod2)[[3]]
```

Plot:
```{r}
ggplot(aes(life_expectancy, tfr, color = dev_region), data = country_ind_2017) + 
  geom_point() + 
  ggtitle("TFR versus life expectancy, by region") + 
  ylab("TFR") + xlab("Life expectancy") + 
  scale_color_brewer(name = "Developed region", palette = "Set1")+
  geom_segment(aes(x = min_max$min[1], xend = min_max$max[1],
                   y = intercept_non_dev+ slope*min_max$min[1], 
                   yend = intercept_non_dev+ slope*min_max$max[1])) +
  geom_segment(aes(x = min_max$min[2], xend = min_max$max[2],
                   y = intercept_dev+ slope*min_max$min[2], 
                   yend = intercept_dev+ slope*min_max$max[2]), color = "blue")
```

# Multinomial

Question of interest: how does infant mortality cause of death vary by race, mother's age and prematurity?

## Data prep

Read in infant data and do some cleaning:


```{r}
d <- read_rds(here("data/infant.RDS")) # this is an RDS file (not csv): use read_rds NOT read_csv

d <- d %>% 
  mutate(neo_death = ifelse(aged<=28, 1, 0),
         cod_group = case_when(
           str_starts(cod, "peri") ~ "perinatal",
           cod %in% c("other", "unknown") ~ "other/unknown",
           cod %in% c("sids", "maltreatment", "infection") ~ "exogenous",
           cod %in% c("resp", "heart") ~ "respiratory/heart",
           TRUE ~ "congenital malformations"
         ),
         preterm = ifelse(gest<37, 1, 0)) %>% 
  filter(gest<99, !is.na(mom_age_group)) # removing missing values

infant <- d %>% select(race, mom_age, gest, preterm, cod_group)
infant
```

Graph by race from lecture:


Get proportions:

```{r}
# get proportion of deaths by race and cause of death.

prop_by_race <- infant %>% 
  group_by(race, cod_group) %>% 
  tally() %>% # counting the number of observations in each race/cause group
  group_by(race) %>% 
  mutate(prop = n/sum(n)) %>% # calculate the proportion of observations in each cause group by race
  mutate(cod_group = fct_reorder(cod_group, -prop)) # plotting purposes: reorder such that the first category has the highest proportion. 
prop_by_race
```
Plotting proportions by race: bar chart

```{r}
p <- ggplot(data = prop_by_race, 
            aes(cod_group, prop, fill  = race)) + 
  geom_bar(stat = "identity", position = 'dodge') + # bar chart
  labs(title = "Proportion of infant deaths by cause", x = "cause", y = "proportion") + 
  theme_bw() + 
  scale_fill_brewer(palette = "Set1")
p
```
Before we run our multinomial regression, we need to make the dataset into a wide format. 

```{r}
infant
```

Change the infant dataset:

- Tally up deaths by race/age/prematurity group
- Switch to wide format so that every cause group has it's own column. 


Making into wide format

```{r}
# count deaths by each group
infant <- infant %>% 
  group_by(race, mom_age, gest, preterm, cod_group) %>% 
  tally(name = "deaths") 
infant

# make this wide:
infant_wide <- infant %>% 
  pivot_wider(names_from = cod_group, values_from = deaths) %>% # pivot wider: makes a dataset wide
  mutate_all(.funs = funs(ifelse(is.na(.), 0, .)))
head(infant_wide)
```

Making the Y variable:

```{r}
# define our outcome variable Y
infant_wide$Y <- as.matrix(infant_wide[,c("perinatal",
                                          "exogenous",
                                          "congenital malformations", 
                                          "respiratory/heart", "other/unknown")])
head(infant_wide$Y)
```

## Regression

```{r}
#install.packages("nnet")
library(nnet)
mod_mn <- multinom(Y ~ race+ mom_age+ preterm, data = infant_wide)
summary(mod_mn)
```

Pull out coefficients:

```{r}
coef(mod_mn)
exp(coef(mod_mn))
```

Pull out standard errors:

```{r}
names(summary(mod_mn))
summary(mod_mn)[["standard.errors"]]
```


Exercise: plot coefficient estimates and standard errors.

Tidy up coefficients and standard errors to get a nice data frame that we can plot:

```{r}
coefs <- as_tibble(coef(mod_mn)) %>% 
  mutate(cod_group = rownames(coef(mod_mn))) %>% 
  pivot_longer(-cod_group, names_to = "covariate", values_to = "coefficient") %>% 
  filter(covariate != "(Intercept)")


ses <- as_tibble(summary(mod_mn)[["standard.errors"]]) %>% 
  mutate(cod_group = rownames(summary(mod_mn)[["standard.errors"]])) %>% 
  pivot_longer(-cod_group, names_to = "covariate", values_to = "se") %>% 
  filter(covariate != "(Intercept)")

coef_se <- coefs %>% 
  left_join(ses)

coef_se
```

Plot:

```{r}
ggplot(data = coef_se, aes(cod_group, coefficient, color = covariate))  +
  geom_point() + 
  geom_errorbar(aes(ymin = coefficient - 2*se, ymax = coefficient + 2*se), width = NA) + 
  coord_flip() + 
  facet_wrap(~covariate, scales = "free_x")+
  theme(legend.position = "none")
```



## Predicted probabilities


Set up a dataframe / tibble that has every combination of our covariates that we're interested in. 
Note: we are calculating predict probabilities for mothers aged 30. 
Note II: Predicted probabilities across race/prematurity will change based on mothers age.

```{r}

# set up our values that we want to get probabilities for.
predict_df <- tibble(race = rep(c("NHW", "NHB"), each = 2),
       mom_age = 30, # fixing mother's age to be 30
       preterm = rep(c(0,1),2)) 

predict_df

# get predicted probabilities : use predict function
predicted_probs <- predict(mod_mn, # model object that we want to make predictions from 
                           newdata = predict_df, # the data that we want to  get predictions for
                           type = 'probs') # return on probability scale

predicted_probs


# tidying up and putting the data and predictions in the one tibble 
preds <- bind_cols(predict_df, as_tibble(predicted_probs))
preds
```

preds is in wide format because every cause has it's own column. 
To plot, need data in long format 

```{r}
preds_long <- preds %>% 
  pivot_longer(`perinatal`:`other/unknown`, 
               names_to = "cod_group", values_to = "probability") %>% 
  mutate(preterm = ifelse(preterm==1, "pre-term", "full-term")) 

preds_long


ggplot(data = preds_long, 
       aes(race, probability, fill = cod_group)) + 
  geom_bar(stat = "identity")+
  facet_grid(~preterm) +
  ggtitle("Predicted probabilities of infant death by race, prematurity and cause\nMothers aged 30")
```

Reorder cause of death:

```{r}
preds_long <- preds_long %>% 
  mutate(cod_group = factor(cod_group, 
                            levels = c("perinatal", "congenital malformations", "exogenous", "respiratory/heart", "other/unknown"))) %>% 
  mutate(cod_group = fct_rev(cod_group))


ggplot(data = preds_long, 
       aes(race, probability, fill = cod_group)) + 
  geom_bar(stat = "identity")+
  facet_grid(~preterm) +
  ggtitle("Predicted probabilities of infant death by race, prematurity and cause\nMothers aged 30")
```









