---
title: 'Week 9: Hierarchical Models'
author: "Monica Alexander"
date: "19/03/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Packages

```{r}
library(tidyverse)
library(lme4)
```


# Read in data

Radon dataset, from Gelman and Hill. 

Household level:

```{r}
d <- read.table(url("http://www.stat.columbia.edu/~gelman/arm/examples/radon/srrs2.dat"), header=T, sep=",")
d
```

Also need info on soil uranium.

County level: 

```{r}
cty <- read.table(url("http://www.stat.columbia.edu/~gelman/arm/examples/radon/cty.dat"), header = T, sep = ",")
cty <- cty %>% mutate(fips = 1000 * stfips + ctfips) %>% dplyr::select(fips, Uppm)

cty
```


Join these together, filter to just be Minnesota and tidy up

```{r}
d_mn <- d %>% 
  filter(state=="MN") %>% # just looking at Minnesota 
  mutate(fips = stfips * 1000 + cntyfips) %>% # creating a new fips variable so i can join the two datasets
  select(fips, county, floor, activity) %>% # selecting the variables that i want (fip, county name, floor, activity (radon))
  left_join(cty) %>% # joining the two datasets together
  mutate(log_uran = log(Uppm), log_activity = log(activity)) %>% # creating some new variables (on log scale)
  mutate(county = str_trim(county)) %>% # tidying up the county variable 
  filter(county!= "") %>% # getting rid of missing county names
  filter(!is.na(log_activity), !is.infinite(log_activity)) # getting rid of missing radon measurements


d_mn
```

look at how many counties

```{r}
d_mn %>% 
  select(county) %>% 
  unique()
```

- y_i is log_activity
- x_i is floor
- u_j is log_uran
- 'county' column tells us the group membership j


# Models using `lm`

## Fit with no pooling (Model 1 in lecture)

$$
y_i \sim N(\alpha_{j[i]}^{\text {no pool}}, \sigma^2_y)
$$
```{r}
mod_nopool <- lm(log_activity~county, data = d_mn)
summary(mod_nopool)
```


## Fit intercept only (model 2 in lecture)

$$
y_i \sim N(\mu, \sigma^2_y)
$$
```{r}
mod_pool <- lm(log_activity ~ 1, data = d_mn)
summary(mod_pool)
```


## Fit with floor as covariate (didn't do in lecture)

$$
y_i \sim N(\beta_0 + \beta_1x_i, \sigma^2_y)
$$ 
This is equivalent to

$$
y_i = \beta_0 + \beta_1x_i + \varepsilon_i
$$

```{r}
mod_floor <- lm(log_activity ~ floor, data = d_mn)
summary(mod_floor)
```



# Models using `lmer`

## Intercept only (Model 3 in lecture)

Model is 

$$
y_i \sim N(\alpha_{j[i]}, \sigma^2_y)
$$

with
$$
\alpha_j \sim N(\mu_{\alpha}, \sigma^2_{\alpha})
$$

Fit using `lmer`


```{r}
mod_hier <- lmer(log_activity ~ (1 | county), data = d_mn) ## (variable | group) # model county intercepts hierarchically
summary(mod_hier)
```

- Look under Random effects --> county variance: $\sigma^2_{\alpha}$
- Look under Fixed effects --> Estimate for the Intercept: $\mu_{\alpha}$



## Adding in covariates at household level:  floor covariate

Model is 
$$
y_i \sim N(\alpha_{j[i]} + \beta x_i, \sigma^2_y)
$$
with

$$
\alpha_j \sim N(\mu_{\alpha}, \sigma^2_{\alpha})
$$

```{r}
mod_hier_floor <- lmer(log_activity ~ floor + (1 | county), data = d_mn)
summary(mod_hier_floor)
```

- $\beta$ estimate is -0.627 (under fixed effects, covariate on floor)
- $\mu_{\alpha}$ estimate (state-level mean = mean of the county intercepts) is 1.48 (under fixed effects, Intercept)
- $\sigma^2_{\alpha}$ estimate (variance of the county intercepts) is 0.13


## Adding in county-level covariate: uranium in soil

Model is

$$
y_i \sim (\alpha_{j[i]} + \beta x_i, \sigma^2_y)
$$
with

$$
\alpha_j \sim N(\gamma_0 + \gamma_1 u_j, \sigma^2_{\alpha})
$$


```{r}
mod_hier_group <- lmer(log_activity ~ floor + log_uran + (1 | county), data = d_mn)
summary(mod_hier_group)
```

- $\beta$ estimate is -0.597
- $\gamma_1$ estimate is 0.754: 1% increase in uranium leads to on average a 0.75% increase in radon
- $\gamma_0$ estimate 1.482 (fixed effect intercept): interpretation: expected log radon for county with log uranium = 0


# Extracting stuff

Now focusing on the last model (hierarchical intercepts, household and county covariates)

## Coefficients ($\beta$, $\gamma_0$ and $\gamma_1$)

Pulling out the overall mean (state mean), the effect of uranium, and the effect of floor


```{r}
coef(summary(mod_hier_group)) 
```

## Extract estimates of county intercepts (alpha's)


```{r}
coef(mod_hier_group)
# Intercept : alpha_j's 
```


Pull out estimates for alpha's

```{r}
res <- coef(mod_hier_group)[[1]]["(Intercept)"]
res
```

Tidy this up:

```{r}
res <- res %>% 
  rename(alpha = `(Intercept)`) %>% # rename column
  mutate(county = rownames(res)) # make a county name column

rownames(res) <- NULL # get rid of rownames

res
```


Pull out standard errors for alpha's and join them to the alphas (horrible)

```{r}
#number of counties
n_counties <- length(unique(d_mn$county))

ses <- attr(ranef(mod_hier_group)[[1]], "postVar")[,,1:n_counties]
df <- res %>% bind_cols(ses) 

colnames(df)[3] <- "se"

df

```

### Plot alpha_j's by county

```{r}

# reorder counties by value of alpha

df <- df %>% 
  mutate(county = fct_reorder(county, alpha))

ggplot(data = df, aes(county, alpha)) +
  geom_point() + 
  coord_flip() +
  geom_errorbar(aes(ymin = alpha - se, ymax = alpha + se))
```

Pull out estimates for gamma0 and gamma 1

```{r}
gamma0 <- coef(summary(mod_hier_group))[1,1]
gamma1 <- coef(summary(mod_hier_group))[3,1]

gamma0
gamma1

```


### Plot expected log radon versus uranium level

Get uranium level for each county

```{r}
urans <- d_mn %>% 
  group_by(county, log_uran) %>% 
  slice(1) %>% 
  dplyr::select(county, log_uran) 
```
Join to alpha estimates 

```{r}
df %>% 
  left_join(urans) %>% 
  filter(county!="") %>%
  mutate(y_hat =  log_uran*gamma1 + alpha)
```


# Questions

Using model with household and county covariates:

- What is expected value of log radon for household in Carlton with basement measure and log uranium = 0?
- What is expected value of log radon for household in Carlton with first floor measure and log uranium = 0?
- What is expected value of log radon for household in Carlton with basement measure and log uranium = 0.3?
