---
title: 'Week 9: Hierarchical Models'
author: "Monica Alexander"
date: "19/03/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Packages

```{r}
library(tidyverse)
library(lme4)
```



# Read in data

Household level:

```{r}
d <- read.table(url("http://www.stat.columbia.edu/~gelman/arm/examples/radon/srrs2.dat"), header=T, sep=",")
```

County level: 

```{r}
cty <- read.table(url("http://www.stat.columbia.edu/~gelman/arm/examples/radon/cty.dat"), header = T, sep = ",")
cty <- cty %>% mutate(fips = 1000 * stfips + ctfips) %>% dplyr::select(fips, Uppm)
```


Join these together, filter to just be Minnesota and tidy up

```{r}
d_mn <- d %>% 
  filter(state=="MN") %>% 
  mutate(fips = stfips * 1000 + cntyfips) %>% 
  dplyr::select(fips, county, floor, activity) %>% 
  left_join(cty) %>% 
  mutate(log_uran = log(Uppm), log_activity = log(activity)) %>% 
  mutate(county = str_trim(county)) %>% 
  filter(county!= "") %>% 
  filter(!is.na(log_activity), !is.infinite(log_activity))
```


# Models using `lm`

- Fit with no pooling
- Fit intercept only
- Fit with floor as covariate

# Models using `lmer`


Intercept only

```{r}
mod_hier <- lmer(log_activity ~ (1 | county), data = d_mn)
mod_hier
```
With floor

```{r}
mod_hier_floor <- lmer(log_activity ~ floor + (1 | county), data = d_mn)
mod_hier_floor
```

With floor and uranium

```{r}
mod_hier_group <- lmer(log_activity ~ floor + log_uran + (1 | county), data = d_mn)
mod_hier_group
```

# Extracting stuff

Coefficients

```{r}
coef(summary(mod_hier_group)) 
```

Note difference between `coef` and `ranef`

```{r}
coef(mod_hier_group)
ranef(mod_hier_group)
```
Pull out estimates for alpha's

```{r}
res <- coef(mod_hier_group)[[1]]["(Intercept)"]
```

Pull out standard errors for alpha's and join them to the alphas (horrible)

```{r}
#number of counties
n_counties <- length(unique(d_mn$county))

ses <- attr(ranef(mod_hier_group)[[1]], "postVar")[,,1:n_counties]
df <- res %>% bind_cols(ses) 
colnames(df) <- c("alpha", "se")
df$county <- rownames(df)
rownames(df) <- NULL

```

Pull out estimates for gamma0 and gamma 1

```{r}
gamma0 <- coef(summary(mod_hier_group))[1,1]
gamma1 <- coef(summary(mod_hier_group))[3,1]

```


# Plotting stuff

## Plot the alphas for each county (county intercepts)

## Plot expected log radon versus uranium level

Get uranium level for each county

```{r}
urans <- d_mn %>% 
  group_by(county, log_uran) %>% 
  slice(1) %>% 
  dplyr::select(county, log_uran) 
```
Join to alpha estimates 

```{r}
df %>% 
  left_join(urans) %>% 
  filter(county!="") %>%
  mutate(y_hat =  log_uran*gamma1 + alpha)
```


# Questions

Using model with household and county covariates:

- What is expected value of log radon for household in Carlton with basement measure and log uranium = 0?
- What is expected value of log radon for household in Carlton with first floor measure and log uranium = 0?
- What is expected value of log radon for household in Carlton with basement measure and log uranium = 0.3?
